---
output: 
  pdf_document: 
    keep_tex: yes
    includes:
        in_header: ~/Downloads/preamble_common.tex
---
# Isaac's stuff

-----


-----

### Scraping
```{r}
library(dplyr)
library(rvest)
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(reshape2)
```



## wnba scraping
```{r}
wilson <- 'https://www.basketball-reference.com/wnba/players/w/wilsoa01w/gamelog/2022/'
wil_doc <- rvest::read_html(wilson)

wil_doc %>%
  rvest::html_elements(., xpath = "//*[(@id = 'div_wnba_pgl_basic')]") %>%
  rvest::html_table() -> wil
wil <- wil[[1]]
head(wil)

#wil2 <- mutate_all(wil, function(x) as.numeric(as.character(x)))
#mean(wil2['PTS'])

#wil$eFG<- (wil['FG'] + (0.5*wil['3P']))/wil['FGA']
#wil$eFG ![Screenshot]('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/images/scraping1')
```



-----

### EDA/Probability

-----

### Baseball


#### WAR comparison (Prob)
Link to WAR explaination: https://www.mlb.com/glossary/advanced-stats/wins-above-replacement

Player X has a projected mean WAR of 3 with standard deviation of 2 and player Y has a projected mean WAR of 1.5 with a standard deviation of 3. Assume projected WAR is normally distributed.
Q: What is the probability that Player X outperforms Player Y?
A: 
We want Pr(X>Y) or Pr(X-Y>0).  
Let Z = X-Y.  
E[Z]=1.5
Var(Z)=5
Pr(Z>0)=1-Pr(Z $\leq$ 0)
```{r}
#Calculate probability Z<=0
pr <- pnorm(0,1.5,sqrt(5))
print(1-pr)
```
The Probability that Player X outperforms Player Y is 0.7488. 



#### Injured Baserunner (Prob)
A runner on first base with 2 out and nobody else on base will attempt to steal second base on the first pitch 70% of the time if he is fully healthy but only 10% of the time if he is playing through an injury. Assume that 80% of the player population is healthy. You see a randomly selected runner not attempt a steal in this situation. 
Q: What is the probability that the runner is playing through an injury?
A: 
From Bayes Theorem:  

Pr(Injury given No Steal) = Pr(No Steal given Injury)\*Pr(Injury)/P(No Steal).  

Pr(No Steal given Injury) = 1 - Pr(Steal given Injury) = 0.9.  

Pr(Injury) = 1- Pr(Healthy) = 0.2.  

Pr(No Steal) = Pr(No Steal given Injury)\*Pr(Injury)+Pr(No Steal given Healthy)\*Pr(Healthy).  

Pr(No Steal) = 0.9\*0.2+0.7\*0.8 = 0.74.  

Therefore Pr(Injury given No Steal) = 0.9\*0.2/0.74 = 0.243.  


#### OPS (EDA)
Q: Using the dataset, plot the leagues average OPS from every year in the data to see the progression.
A:
```{r}
mlb = read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/mlb_team_stats_history.csv')
head(mlb)

# make new variables
mlb=mutate(mlb,SLG=(X1B+2*X2B+3*X3B+4*HR)/(AB))
mlb=mutate(mlb,OBP=(H+BB+HBP)/(AB+BB+HBP+SF))
mlb=mutate(mlb,OPS=OBP+SLG)

# get avg ops
summarize(mlb, Average = mean(OPS,na.rm=T))

# get avg ops by year
group_by(mlb, yearID)%>%
summarize(Average = mean(OPS, na.rm=T))

group_by(mlb, yearID)%>%
summarize(Average = mean(OPS, na.rm=T))%>%View

#create new dataset
mlbYr=group_by(mlb, yearID)%>%
summarize(Average = mean(OPS, na.rm=T))

#plot it
ggplot(mlbYr, aes(x=yearID, y= Average))+geom_point()
```
Followup Q: What would cause the data to peak around the year 2000?
A: PED's


#### Run Variance (Probability)
\begin{center}
\begin{tabular}{ c c c c c }
 Runs Scored & Probability \\ 
 0 & 0.55 \\  
 1 & 0.25 \\
 2 & 0.15 \\
 3 & 0.05
\end{tabular}
\end{center} 


Q: Using the probability table provided, calculate the variance for runs scored in an inning
A:
$E(X)=1*0.25+2*0.15+3*0.05=0.7$  
$E(X^2)=1*0.25+4*0.15+9*0.05=1.3$  
$Var(X)=E(X^2)-E(X)=1.3-0.7=0.6$


-----

### Tennis
Link for brief explanation of tennis scoring: https://www.sportingnews.com/us/tennis/news/tennis-scoring-explained-rules-system-points-terms/7uzp2evdhbd11obdd59p3p1cx


#### Probability of Winning a Game (Prob)
The formula for the probability of a tennis player winning a game (from Analyzing Wimbledon) is given by
$\frac{p^4*(-8*p^3+28*p^2-34*p+15)}{p^2+(1-p)^2}$ where $p$ is the probability of a player winning their service point.
Q: If a player wins their service points 62% of the time, what is the probability they win the game?
A:
```{r}
p <- 0.62
pr_game <- (p^4*(-8*p^3+28*p^2-34*p+15))/(p^2+(1-p)^2)
pr_game
```


#### Graph Example of Probability of Winning Point vs Probability of Winning Game (Prob)
```{r}
game <- c(0)
pr <- 1:100
for(x in pr) {
  p <- pr/100
  pr_game <- (p^4*(-8*p^3+28*p^2-34*p+15))/(p^2+(1-p)^2)
  game <- c(game,pr_game)
}
game[1]
game <- game[2:101]
game[1]
df <- do.call(rbind, Map(data.frame, point_pr=pr, game_pr=game))
ggplot(df, aes(x=point_pr, y=game_pr)) +
  geom_point()+xlab('Probability of Winning a Service Point')+ylab('Probability of Winning a Game')
```


### WNBA Scores (EDA)
Q: What is the difference in PPG for a winning team at home vs a winning team away?
A:
```{r}
wnba=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/WNBA_Games2019_Scores.csv')
head(wnba)

group_by(wnba, Winner)%>%
  summarize(Count=n())%>%
  mutate(Percent=Count/sum(Count))

group_by(wnba, Winner)%>%
  summarize(Average=mean(PTSwin,na.rm=T),sd=sd(PTSwin,na.rm=T))

84.822-83.787
```
A home team winner scores on average 1.035 PPG more than an away team winner.


### NFL
```{r}
nfl=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/nfl_pbp.csv')
nfl2 <- select(nfl, c('Date','GameID','qtr','down','time','yrdline100','ydstogo','Yards.Gained','Touchdown','PlayType','FieldGoalResult','FieldGoalDistance','ScoreDiff','Season'))
head(nfl2)
```

#### 4th Down Analysis (EDA)
Q: Using NFL Play by Play data, what percentage of the time do coaches choose to go for it on 4th down? And what percentage of 4th down attempts are successful?
A:
```{r}
# add indicator column for successful first down attempt
nfl2 <- nfl2 %>%
  mutate(FirstDown = case_when(
    ydstogo < Yards.Gained ~ 1,
    ydstogo > Yards.Gained ~ 0
    ))
# filter by only plays on 4th down
down4 = filter(nfl2, nfl2['down']==4)

#see what play types are run on fourth down and remove the noise
group_by(down4,PlayType) %>%
  summarize(Count=n())%>%
  mutate(Percentage=Count/sum(Count))
down4 = filter(down4, down4['PlayType']!='No Play' || down4['PlayType']!= 'QB Kneel' || down4['PlayType']!= 'Timeout')

# add indicator column for going for it on 4th
down4 <- down4 %>% 
  mutate(GoForIt = case_when(
    PlayType == 'Pass' ~ 1,
    PlayType == 'Run' ~ 1,
    PlayType == 'Sack' ~ 1,
    PlayType == 'Field Goal' ~ 0, 
    PlayType == 'Punt' ~ 0
  ))
# get percentage of 4th downs are gone for
group_by(down4,GoForIt) %>%
  summarize(Count=n())%>%
  mutate(Percentage=Count/sum(Count))

# get percentage of successful attempted 4th downs 
down4 %>%
  filter(down4['GoForIt']==1) %>%
  group_by(FirstDown) %>%
    summarize(Count=n())%>%
    mutate(Percentage=Count/sum(Count))  
  
```
11% of 4th downs are gone for and 40% of those are successful, regardless of how many yards to go there are





#### Probability of Outcome based on Field Position Graph
```{r}
nfl=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/nfl_pbp.csv')
nfl2 <- select(nfl, c('yrdline100','ydstogo','No_Score_Prob','Field_Goal_Prob','Touchdown_Prob'))
nfl2 %>%
  na.omit() -> nfl2
```

```{r}
nfl2 %>%
  group_by(yrdline100) %>%
    summarize(mean(Touchdown_Prob, na.rm=T)) -> td_prob
nfl2 %>%
  group_by(yrdline100) %>%
    summarize(mean(Field_Goal_Prob, na.rm=T)) -> fg_prob
nfl2 %>%
  group_by(yrdline100) %>%
    summarize(mean(No_Score_Prob, na.rm=T)) -> no_prob
x <- c('yrdline100','probability', 'Outcome')
colnames(td_prob) <- x
ind <- data.frame(ncol = 1, nrow=nrow(td_prob))
ind= 'TD'
td_prob <- cbind(td_prob,ind)
ind2 <- data.frame(ncol=1, nrow=nrow(fg_prob))
ind2='FG'
fg_prob <- cbind(fg_prob, ind2)
colnames(td_prob) <- x
colnames(fg_prob) <- x
prob_df <- rbind(td_prob, fg_prob)
colnames(prob_df) <- x
ggplot(prob_df, aes(yrdline100,probability, col=Outcome)) + 
  geom_point(size=0.5)+geom_smooth()+xlab('Distance to Endzone')+ylab('Probability')+ggtitle('Probability of Outcome based on Field Position')

#+facet_wrap('ydstogo')
```
Q: Why does the probability of scoring a field goal get lower as a team is within 10 yards of the endzone?

A: When a team is close to the endzone, they probability of scoring a touchdown goes way up so teams are less likely to attempt field goals since the expected value of attempting a touchdown is higher than the expected value of attempting a field goal.


#### IN PROGRESS Should they go for it? (Based on historical data)
```{r}
yrd = 45
togo = 3


nfl=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/nfl_pbp.csv')
nfl2 <- select(nfl, c('down','yrdline100','ydstogo','Yards.Gained','Touchdown','PlayType','FieldGoalResult','FieldGoalDistance'))
```

```{r}
nfl2 <- nfl2 %>%
  mutate(FirstDown = case_when(
    ydstogo < Yards.Gained ~ 1,
    ydstogo > Yards.Gained ~ 0
    ))
# filter by only plays on 4th down
down4 = filter(nfl2, nfl2['down']==4)
down4 <- down4 %>% 
  filter(PlayType!= 'Punt') %>%
  filter(PlayType!= 'No Play') %>%
  filter(PlayType!= 'QB Kneel') %>%
  filter(PlayType!= 'Timeout') %>%
  mutate(GoForIt = case_when(
    PlayType == 'Pass' ~ 1,
    PlayType == 'Run' ~ 1,
    PlayType == 'Sack' ~ 1,
    PlayType == 'Field Goal' ~ 0 
  ))

down4 <- down4 %>%
  mutate(Success = case_when(
    FieldGoalResult == 'Good' ~ 1,
    FirstDown == 1 ~ 1,
    TRUE ~ 0
  ))
down5 <- down4 %>%
  group_by(yrdline100) %>%
        group_by(Success) %>%
          summarize(Count = n()) %>%
          mutate(Percentage=Count/sum(Count))

head(down5)
'
down5<- down4 %>%
  group_by(yrdline100) %>%
    group_by(ydstogo) %>%
      group_by(GoForIt) %>%
        group_by(Success) %>%
          summarize(Count = n()) %>%
          mutate(Percentage=Count/sum(Count))
'
# create list of plays (list)
# for i in 0-100:
#   count = 0
#   if yrdline100 = i:
#       count = count+1
#       list[i][count] = c(ydstogo, success, goforit)

# group by yards to go

# create list of probabilities (list2)
#create final list (list_f)
# for i in 0-100:
#   create sublist (sublist)
#   lst = list2[i]
#   for i in length(lst):
#       add lst[i][]


## want to find every play with same situation and group by playtype and find success rate for going for it or kicking field goal
```


#### Football Sample Space (Probability)
A sample space contains all possible outcomes. An american football game can either end with a win (W), loss (L) or a tie (T) which means our sample space is $\Omega =  \{W,L,T \}$ and an event, $E$ would be one of the possible outcomes. If a team wins the game, the event for that game would be $E=\{W \}$ or if we want the event of the 2021 CSU football season, it would be $E=\{ L, L, W, L, W, W, L, L, L, L, L, L \}$.

------------------------

## Gambling

### Sports Betting Bankroll Management
To prevent problematic gambling many people use bankrolls, or money set aside with the sole purpose of gambling. It is often suggested that people partaking in sports betting set aside an amount they are comfortable gambling with and betting 1-5% of that per play, especially considering minimum bets for online sportsbooks are often less than $1. One of the big risks in gambling is chasing losses so a popular strategy is called flat betting, where you bet the same amount on every game to help minimize losses. In addition, parlays can have incredibly attractive payoffs but high reward comes with high risk so it can be quite difficult to find long term profit. 


### Hold Percentage/Breaking Even 
In a perfect world, if a baseball game has probability 0.5 of either team winning, the odds would be +100 (or -100), 50% of bettors would be on one side, 50% on the other, and at the end of the game half of the bettors would double their money and half of them would lose their money. Unfortunately, this would mean that the casinos make no money so to combat this they introduce whats called a hold percentage. Essentially sportsbooks will give you slightly worse odds on bets in order to make money. In this example, when both sides have probability 0.5, the offered lines may both be -110 that way when one team wins, half the bettors win slightly less than double the money (bet 110 dollars to win 100) and the sportsbook collects the rest. As a result, winning percentages have to be higher than expected to show a profit. For -110 odds (the Vegas equivalent of probability = 0.5), a bettor must win at a rate of 52.23% of the time in order to show a profit. Similarly you can convert any American odds into implied probability to get a breakeven percentage, since if you have a win rate higher than the implied probability, the expected value of your bet is positive. 

Q: The given odds for the Avs to win the 2022 Stanley Cup are +500 and you've concluded that the bet has a 15% chance of happening. Is it worth making this bet?

A: The implied probability on a +500 bet is 16.66%, therefore if the bet has a 15% chance of happening, it is not a good bet as the expected value is negative but lets look at a simulation.
```{r}
# lets say we start with $1 and we want to bet $1
dollars = 1
set.seed(1)
# we want to simulate the bet 100,000 times
n.sims <- 100000
for(i in 1:n.sims){
# Simulate whether the Avs win based on our probability of 15%
  win <- sum(rbinom(1,1,0.15))
  # change our dollar amount based off the odds
  if(win == 1){
    dollars = dollars + 5
  }
  if(win == 0){
    dollars = dollars - 1
  }
}
print(dollars)
```
After 100,000 bets, we would be down 9,573 dollars just off of 1 dollar bets.


### Gamblers Fallacy
The Gambler's Fallacy is the common misbelief that if independent events fail to happen, theyre more likely to happen in the future. For example, say we are playing Roulette and betting on whether the ball lands on black or red. Theres roughly a 50% chance of either outcome, yet if we somehow get 10 reds in a row, the natural inclination is to assume a black is "due" and must come soon. However, both outcomes still have just a 50% chance of happening regardless of the history. This can be seen in the sports world as well, especially when talking about something like the 'hot hand' in basketball or hitting streaks in baseball, although sports does differ from conventional gambling as recent performance in sports can be useful for predictive purposes. 

### Kelly Criterion
While flat betting is a common and effective bet sizing strategy, a more advanced technique is called the Kelly Criterion. The Kelly Criterion adjusts each bet size to the specific bet based on bankroll size, given odds, and the predicted probability that the bettor gives the outcome. The formula for calculating the bet size is $f=p-\frac{1-p}{b}$ where $f$ is the fraction of your current bankroll you should wager, $p$ is the probability the bettor gives themselves of winning the bet, and $b$ is the proportion of the bet you win back (+200 odds pays 2 to 1 therefore the proportion is 2.0). This is certainly more complicated than flat betting (especially in large volumes), but it has been proven to provide theoretically optimal bet sizes.



#### SANDBOX (Can be ignored)
```{r}

nfl=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/nfl_pbp.csv')
nfl2 <- select(nfl, c('down','yrdline100','ydstogo','Touchdown','PlayType','FieldGoalResult','FieldGoalDistance'))
head(nfl2)


## for each (grouping of 10 yards?) we get all the 4th down playtypes of fieldgoal and normal plays and see probability of success for each option, for each group
# filter by only plays on 4th down
down4 = filter(nfl2, nfl2['down']==4)

#see what play types are run on first down and remove the noise
group_by(down4,PlayType) %>%
  summarize(Count=n())%>%
  mutate(Percentage=Count/sum(Count))
down3 = filter(down4, down4['PlayType']!='No Play' || down4['PlayType']!= 'QB Kneel' || down4['PlayType']!= 'Timeout' || down4['PlayType']!= 'Punt')
head(down3)
## create 10 grouping dataframes (99-90,89-80,79-70,69-60,59-50,49-40,39-30,29-20,19-10,9-0)
x <- c("playtype", "outcome")
df99_90 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df99_90) <- x
df89_80 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df89_80) <- x
df79_70 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df79_70) <- x
df69_60 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df69_60) <- x
df59_50 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df59_50) <- x
df49_40 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df49_40) <- x
df39_30 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df39_30) <- x
df29_20 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df29_20) <- x
df19_10 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df19_10) <- x
df09_00 <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df09_00) <- x



## create final dataframe with 3 columns: distance to goal, field goal probability, td probability
#df_prob <- data.frame(matrix(ncol = 3, nrow = 0))
#y = c('distance_to_goal','fg_prob','td_prob')
#colnames(df_prob) <- x
#down3 %>%
#  group_split(yrdline100)

down3 %>%
  mutate(distance = case_when(down3['yrdline100']<100 & down3['yrdline100'] >89 ~ 90,
                           down3['yrdline100']<90 & down3['yrdline100'] >79 ~ 80,
                           down3['yrdline100']<80 & down3['yrdline100'] >69 ~ 70,
                           down3['yrdline100']<70 & down3['yrdline100'] >59 ~ 60,
                           down3['yrdline100']<60 & down3['yrdline100'] >49 ~ 50,
                           down3['yrdline100']<50 & down3['yrdline100'] >39 ~ 40,
                           down3['yrdline100']<40 & down3['yrdline100'] >29 ~ 30,
                           down3['yrdline100']<30 & down3['yrdline100'] >19 ~ 20,
                           down3['yrdline100']<20 & down3['yrdline100'] >9 ~ 10,
                           down3['yrdline100']<10 ~ 0
    )) -> down3
down3 %>%
  group_split(distance) -> yrd_df

df99_90 <- yrd_df[[10]]
df89_80 <- yrd_df[[9]]
df79_70 <- yrd_df[[8]]
df69_60 <- yrd_df[[7]]
df59_50 <- yrd_df[[6]]
df49_40 <- yrd_df[[5]]
df39_30 <- yrd_df[[4]]
df29_20 <- yrd_df[[3]]
df19_10 <- yrd_df[[2]]
df09_00 <- yrd_df[[1]]
df09_00 %>%
  filter(df09_00['PlayType']=='Field Goal') %>%
  group_by(FieldGoalResult) %>%
    summarize(Count=n())%>%
    mutate(Percentage=Count/sum(Count))
df09_00 %>%
  filter(df09_00['PlayType']!='Field Goal' && df09_00['PlayType']!='Punt') %>%
  group_by(Touchdown) %>%
    summarize(Count=n())%>%
    mutate(Percentage=Count/sum(Count))

```

