---
output: 
  pdf_document: 
    keep_tex: yes
    includes:
        in_header: ~/Downloads/preamble_common.tex
---
# Isaac's stuff

-----


-----

### Scraping
```{r}
library(dplyr)
library(rvest)
library(tidyverse)
library(kableExtra)

```



## wnba scraping
```{r}
wilson <- 'https://www.basketball-reference.com/wnba/players/w/wilsoa01w/gamelog/2022/'
wil_doc <- rvest::read_html(wilson)

wil_doc %>%
  rvest::html_elements(., xpath = "//*[(@id = 'div_wnba_pgl_basic')]") %>%
  rvest::html_table() -> wil
wil <- wil[[1]]
head(wil)

#wil2 <- mutate_all(wil, function(x) as.numeric(as.character(x)))
#mean(wil2['PTS'])

#wil$eFG<- (wil['FG'] + (0.5*wil['3P']))/wil['FGA']
#wil$eFG ![Screenshot]('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/images/scraping1')
```



-----

### EDA/Probability

-----

### Baseball


#### WAR comparison (Prob)
Link to WAR explaination: https://www.mlb.com/glossary/advanced-stats/wins-above-replacement

Player X has a projected mean WAR of 3 with standard deviation of 2 and player Y has a projected mean WAR of 1.5 with a standard deviation of 3. Assume projected WAR is normally distributed.
Q: What is the probability that Player X outperforms Player Y?
A: 
We want Pr(X>Y) or Pr(X-Y>0).  
Let Z = X-Y.  
E[Z]=1.5
Var(Z)=5
Pr(Z>0)=1-Pr(Z $\leq$ 0)
```{r}
#Calculate probability Z<=0
pr <- pnorm(0,1.5,sqrt(5))
print(1-pr)
```
The Probability that Player X outperforms Player Y is 0.7488. 



#### Injured Baserunner (Prob)
A runner on first base with 2 out and nobody else on base will attempt to steal second base on the first pitch 70% of the time if he is fully healthy but only 10% of the time if he is playing through an injury. Assume that 80% of the player population is healthy. You see a randomly selected runner not attempt a steal in this situation. 
Q: What is the probability that the runner is playing through an injury?
A: 
From Bayes Theorem:  

Pr(Injury given No Steal) = Pr(No Steal given Injury)\*Pr(Injury)/P(No Steal).  

Pr(No Steal given Injury) = 1 - Pr(Steal given Injury) = 0.9.  

Pr(Injury) = 1- Pr(Healthy) = 0.2.  

Pr(No Steal) = Pr(No Steal given Injury)\*Pr(Injury)+Pr(No Steal given Healthy)\*Pr(Healthy).  

Pr(No Steal) = 0.9\*0.2+0.7\*0.8 = 0.74.  

Therefore Pr(Injury given No Steal) = 0.9\*0.2/0.74 = 0.243.  


#### OPS (EDA)
Q: Using the dataset, plot the leagues average OPS from every year in the data to see the progression.
A:
```{r}
mlb = read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/mlb_team_stats_history.csv')
head(mlb)

# make new variables
mlb=mutate(mlb,SLG=(X1B+2*X2B+3*X3B+4*HR)/(AB))
mlb=mutate(mlb,OBP=(H+BB+HBP)/(AB+BB+HBP+SF))
mlb=mutate(mlb,OPS=OBP+SLG)

# get avg ops
summarize(mlb, Average = mean(OPS,na.rm=T))

# get avg ops by year
group_by(mlb, yearID)%>%
summarize(Average = mean(OPS, na.rm=T))

group_by(mlb, yearID)%>%
summarize(Average = mean(OPS, na.rm=T))%>%View

#create new dataset
mlbYr=group_by(mlb, yearID)%>%
summarize(Average = mean(OPS, na.rm=T))

#plot it
ggplot(mlbYr, aes(x=yearID, y= Average))+geom_point()
```
Followup Q: What would cause the data to peak around the year 2000?
A: PED's


#### Run Variance (Probability)
\begin{center}
\begin{tabular}{ c c c c c }
 Runs Scored & Probability \\ 
 0 & 0.55 \\  
 1 & 0.25 \\
 2 & 0.15 \\
 3 & 0.05
\end{tabular}
\end{center} 


Q: Using the probability table provided, calculate the variance for runs scored in an inning
A:
$E(X)=1*0.25+2*0.15+3*0.05=0.7$  
$E(X^2)=1*0.25+4*0.15+9*0.05=1.3$  
$Var(X)=E(X^2)-E(X)=1.3-0.7=0.6$


-----

### Tennis
Link for brief explanation of tennis scoring: https://www.sportingnews.com/us/tennis/news/tennis-scoring-explained-rules-system-points-terms/7uzp2evdhbd11obdd59p3p1cx


#### Probability of Winning a Game (Prob)
The formula for the probability of a tennis player winning a game (from Analyzing Wimbledon) is given by
$\frac{p^4*(-8*p^3+28*p^2-34*p+15)}{p^2+(1-p)^2}$ where $p$ is the probability of a player winning their service point.
Q: If a player wins their service points 62% of the time, what is the probability they win the game?
A:
```{r}
p <- 0.62
pr_game <- (p^4*(-8*p^3+28*p^2-34*p+15))/(p^2+(1-p)^2)
pr_game
```


#### Graph Example of Probability of Winning Point vs Probability of Winning Game (Prob)
```{r}
game <- c(0)
pr <- 1:100
for(x in pr) {
  p <- pr/100
  pr_game <- (p^4*(-8*p^3+28*p^2-34*p+15))/(p^2+(1-p)^2)
  game <- c(game,pr_game)
}
game[1]
game <- game[2:101]
game[1]
df <- do.call(rbind, Map(data.frame, point_pr=pr, game_pr=game))
ggplot(df, aes(x=point_pr, y=game_pr)) +
  geom_point()+xlab('Probability of Winning a Service Point')+ylab('Probability of Winning a Game')
```


### WNBA Scores (EDA)
Q: What is the difference in PPG for a winning team at home vs a winning team away?
A:
```{r}
wnba=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/WNBA_Games2019_Scores.csv')
head(wnba)

group_by(wnba, Winner)%>%
  summarize(Count=n())%>%
  mutate(Percent=Count/sum(Count))

group_by(wnba, Winner)%>%
  summarize(Average=mean(PTSwin,na.rm=T),sd=sd(PTSwin,na.rm=T))

84.822-83.787
```
A home team winner scores on average 1.035 PPG more than an away team winner.


### NFL
```{r}
nfl=read.csv('~/Google Drive/My Drive/Sports Analytics/SportsAnalyticsBook/data/nfl_pbp.csv')
nfl2 <- select(nfl, c('Date','GameID','qtr','down','time','yrdline100','ydstogo','Yards.Gained','Touchdown','PlayType','FieldGoalResult','FieldGoalDistance','ScoreDiff','Season'))
head(nfl2)
```

#### 4th Down Analysis (EDA)
Q: Using NFL Play by Play data, what percentage of the time do coaches choose to go for it on 4th down? And what percentage of 4th down attempts are successful?
A:
```{r}
# add indicator column for successful first down attempt
nfl2 <- nfl2 %>%
  mutate(FirstDown = case_when(
    ydstogo < Yards.Gained ~ 1,
    ydstogo > Yards.Gained ~ 0
    ))
# filter by only plays on 4th down
down4 = filter(nfl2, nfl2['down']==4)

#see what play types are run on first down and remove the noise
group_by(down4,PlayType) %>%
  summarize(Count=n())%>%
  mutate(Percentage=Count/sum(Count))
down4 = filter(down4, down4['PlayType']!='No Play' || down4['PlayType']!= 'QB Kneel' || down4['PlayType']!= 'Timeout')

# add indicator column for going for it on 4th
down4 <- down4 %>% 
  mutate(GoForIt = case_when(
    PlayType == 'Pass' ~ 1,
    PlayType == 'Run' ~ 1,
    PlayType == 'Sack' ~ 1,
    PlayType == 'Field Goal' ~ 0, 
    PlayType == 'Punt' ~ 0
  ))
# get percentage of 4th downs are gone for
group_by(down4,GoForIt) %>%
  summarize(Count=n())%>%
  mutate(Percentage=Count/sum(Count))

# get percentage of successful attempted 4th downs 
down4 %>%
  filter(down4['GoForIt']==1) %>%
  group_by(FirstDown) %>%
    summarize(Count=n())%>%
    mutate(Percentage=Count/sum(Count))  
  
```
11% of 4th downs are gone for and 40% of those are successful, regardless of how many yards to go there are



#### Football Sample Space (Probability)
A sample space contains all possible outcomes. An american football game can either end with a win (W), loss (L) or a tie (T) which means our sample space is $\Omega =  \{W,L,T \}$ and an event, $E$ would be one of the possible outcomes. If a team wins the game, the event for that game would be $E=\{W \}$ or if we want the event of the 2021 CSU football season, it would be $E=\{ L, L, W, L, W, W, L, L, L, L, L, L \}$.

