---
output:
  pdf_document: default
  html_document: default
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(formatR)
library(tidyverse)
library(kableExtra)

# Set so that long lines in R will be wrapped:
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=FALSE)
knitr::opts_chunk$set(cache=TRUE)

# kable table global setup
kt <- function(data) {
  knitr::kable(data, digits=3, align=c("c")) %>% kable_styling(bootstrap_options='striped', latex_options='HOLD_position', full_width = F, position = "center")
}
```

# Preliminaries

## Getting Started With R

### Installing R

For this class, you will be using R Studio to complete statistical analyses on your computer. 

To begin using R Studio, you will need to install "R" first and then install "R Studio" on your computer.

***Step 1: Download R*** \
(a) Visit [https://www.r-project.org/](https://www.r-project.org/) \
(b) Click **CRAN** under **Download**
(c) Select any of the mirrors \
(d) Click the appropriate link for your type of system (Mac, Windows, Linux) \
(e) Download R on this next page. \
(For Windows, this will say **install R for the first time**. For Mac, this will be under **Latest release** and will be something like **R-4.1.0.pkg** -- the numbers may differ depending on the most recent version) \
(f) Install R on your computer \

***Step 2: Download R Studio*** \
(a) Visit [https://www.rstudio.com/products/rstudio/download/#download](https://www.rstudio.com/products/rstudio/download/#download) \
(b) Click to download \
(c) Install R Studio on your computer \

\newpage

***Step 3: Verify R Studio is working*** \
(a)  Open R Studio \
(b) Let's enter a small dataset and calculate the average to make sure everything is working correctly. \
(c) In the console, type in the following dataset of Sammy Sosa's season home run totals from 1998--2002: \

```{r}
sosa.HR <- c(66,63,50,64,49)
```

(d) In the console, calculate the average season home run total for Sammy Sosa between 1998--2002: \

```{r}
mean(sosa.HR)
```

\noindent (e) Did you find Slammin' Sammy's average home run total from 1998--2002 was 58.4? If so, you should be set up correctly!

\newpage

### Some R Basics

For the following examples, let's consider Peyton Manning's career with the Denver Broncos. In his four seasons with the Broncos, Manning's passing yard totals were: 4659, 5477, 4727, 2249. Let's enter this data into R. To enter a vector of data, use the **c()** function.

```{r}
peyton <- c(4659, 5477, 4727, 2249)
```

To look at the data you just put in the variable *peyton*, type *peyton* into the console and press enter.

```{r}
peyton
```

Some basic function for calculating summary statistics include **summary**, **mean()**, **median()**, **var()**, and **sd()**.

```{r}
summary(peyton)
mean(peyton)
sd(peyton)
```

R allows you to install additional packages (collections of functions) that aren't offered in the base version of R. To install a package, use **install.packages(<name_of_package>)** and to load a package, use **library(<name_of_package>)**.

One package that we will use frequently is **tidyverse**. This package includes several other packages and functions such as **ggplot** (plotting function), **dplyr** (data manipulation package), and **stringr** (string manipulation package).

```{r,message=F,include=F}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
```

```{r,message=F,eval=F}
install.packages("tidyverse")
library("tidyverse")
```

You will also need to know how to load datasets from files. For this class, we will typically provide data files is .csv format.

\newpage

Here is how to load a file:
```{r load, message=F}
# load readr package and load example dataset
NFL_2021_Team_Passing <- read_csv("data/NFL_2021_Team_Passing.csv")

# we can look at the header (first few entries) using "head()"
head(NFL_2021_Team_Passing)
```

\newpage


## Which Sports?

Which sports will we be analyzing in this class?

\vfill


## Challenges

What kind of challenges could we encounter dealing with sports data?

\vfill



## Cautionary Notes

What should we be careful of when analyzing sports data?

\vfill

## Ethical Concerns

What kinds of ethical concerns should we consider when dealing with sports data?

\vfill
