# Exploratory Data Analysis

-----

## Installing R

For this class, you will be using R Studio to complete statistical analyses on your computer. 

To begin using R Studio, you will need to install "R" first and then install "R Studio" on your computer.

***Step 1: Download R*** \
(a) Visit [https://www.r-project.org/](https://www.r-project.org/) \
(b) Click **CRAN** under **Download**
(c) Select any of the mirrors \
(d) Click the appropriate link for your type of system (Mac, Windows, Linux) \
(e) Download R on this next page. \
(For Windows, this will say **install R for the first time**. For Mac, this will be under **Latest release** and will be something like **R-4.1.0.pkg** -- the numbers may differ depending on the most recent version) \
(f) Install R on your computer \
       
***Step 2: Download R Studio*** \
(a) Visit [https://www.rstudio.com/products/rstudio/download/#download](https://www.rstudio.com/products/rstudio/download/#download) \
(b) Click to download \
(c) Install R Studio on your computer \

***Step 3: Verify R Studio is working*** \
(a)  Open R Studio \
(b) Let's enter a small dataset and calculate the average to make sure everything is working correctly. \
(c) In the console, type in the following dataset of Sammy Sosa's season home run totals from 1998--2002: \

```{r}
sosa.HR <- c(66,63,50,64,49)
```

(d) In the console, calculate the average season home run total for Sammy Sosa between 1998--2002: \

```{r}
mean(sosa.HR)
```

\noindent (e) Did you find Slammin' Sammy's average home run total was 58.4? If so, you should be set up correctly!

-----

## Getting Started With R

For the following examples, let's consider Peyton Manning's career with the Denver Broncos. In his four seasons with the Broncos, Manning's passing yard totals were: 4659, 5477, 4727, 2249. Let's enter this data into R. To enter a vector of data, use the **c()** function.

```{r}
peyton <- c(4659, 5477, 4727, 2249)
```

To look at the data you just put in the variable *peyton*, type *peyton* into the console and press enter.

```{r}
peyton
```

Some basic function for calculating summary statistics include **mean()**, **median()**, **var()**, and **sd()**.

```{r}
mean(peyton)
median(peyton)
var(peyton)
sd(peyton)
```

R allows you to install additional packages (or functions) that aren't offered in the base version of R. To install a package, use **install.packages(<name_of_package>)** and to load a package, use **library(<name_of_package>)**.

One package that we will use frequently is **tidyverse**. This package includes several other packages and functions such as **ggplot** (plotting function), **dplyr** (data manipulation package), and **stringr** (string manipulation package).

```{r,message=F,include=F}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
```

```{r,message=F,eval=F}
install.packages("tidyverse")
library("tidyverse")
```

-----

## Definitions

::: {.definition}
A ***population*** is a well-defined complete collection of objects.
:::

::: {.definition}
A ***sample*** is a subset of the population.
:::

::: {.example}
Suppose we are interested in studying Peyton's Manning's season passing yards totals. What would be the population and what is one possible sample?
::: 

\
\
\
\
\

::: {.example}
Suppose we are interested in studying Peyton's Manning's season passing yards totals. What would be the population and what is one possible sample?
::: 

\
\
\
\
\

::: {.definition}
***Quantitative data*** is numeric data or numbers. It can be broken into two further categories: discrete and continuous data. 
:::

::: {.definition}
***Discrete data*** is quantitative data with a finite or countably infinite number of values. 
:::

::: {.definition}
***Continuous data*** is quantitative data with an uncountably infinite number of values or data taken from an interval.
:::

::: {.example}
What are possible discrete and continuous data associated with Peyton Manning?
:::

\
\
\
\
\

::: {.definition}
***Qualitative data*** refers to names, categories, or descriptions. It can also be broken down into
two further categories, nominal data and ordinal data.
:::

::: {.definition}
***Nominal data*** is qualitative data with no natural ordering.
:::

::: {.definition}
***Ordinal data*** is qualitative data with a natural ordering.
:::

::: {.example}
What are possible nominal and ordinal data associated with Peyton Manning?
:::

\
\
\
\
\

-----


## Descriptive Statistics

While we will learn about some descriptive statistics that are unique to specific sports, there are some descriptive statistics that are frequently used in many applications.

To begin, let $x_1, x_2, \ldots, x_n$ represent a sample of size $n$, where $x_i$ is the $i^\text{th}$ value in the dataset.


::: {.definition}
The ***sum*** of the data values is given by: $\sum_{i=1}^n x_i = x_1 + x_2 + \ldots + x_n$
:::

::: {.definition}
The ***sample mean*** of the dataset is given by $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i$
:::

::: {.example}
Recall that Peyton Manning's season passing yards total while with the Broncos were: 4659, 5477, 4727, 2249. Calculate the sample mean of these values.
:::

\
\
\
\
\


-----

[NFL Team Passing Statistics, 2021](https://www.pro-football-reference.com/years/2021/index.htm#passing)

## Baseball

## Football

## Basketball

## Soccer

## Volleyball

## Hockey