# Exploratory Data Analysis

-----

## Getting Started With R

### Installing R

For this class, you will be using R Studio to complete statistical analyses on your computer. 

To begin using R Studio, you will need to install "R" first and then install "R Studio" on your computer.

***Step 1: Download R*** \
(a) Visit [https://www.r-project.org/](https://www.r-project.org/) \
(b) Click **CRAN** under **Download**
(c) Select any of the mirrors \
(d) Click the appropriate link for your type of system (Mac, Windows, Linux) \
(e) Download R on this next page. \
(For Windows, this will say **install R for the first time**. For Mac, this will be under **Latest release** and will be something like **R-4.1.0.pkg** -- the numbers may differ depending on the most recent version) \
(f) Install R on your computer \
       
***Step 2: Download R Studio*** \
(a) Visit [https://www.rstudio.com/products/rstudio/download/#download](https://www.rstudio.com/products/rstudio/download/#download) \
(b) Click to download \
(c) Install R Studio on your computer \

***Step 3: Verify R Studio is working*** \
(a)  Open R Studio \
(b) Let's enter a small dataset and calculate the average to make sure everything is working correctly. \
(c) In the console, type in the following dataset of Sammy Sosa's season home run totals from 1998--2002: \

```{r}
sosa.HR <- c(66,63,50,64,49)
```

(d) In the console, calculate the average season home run total for Sammy Sosa between 1998--2002: \

```{r}
mean(sosa.HR)
```

\noindent (e) Did you find Slammin' Sammy's average home run total from 1998--2002 was 58.4? If so, you should be set up correctly!

-----

### Some R Basics

For the following examples, let's consider Peyton Manning's career with the Denver Broncos. In his four seasons with the Broncos, Manning's passing yard totals were: 4659, 5477, 4727, 2249. Let's enter this data into R. To enter a vector of data, use the **c()** function.

```{r}
peyton <- c(4659, 5477, 4727, 2249)
```

To look at the data you just put in the variable *peyton*, type *peyton* into the console and press enter.

```{r}
peyton
```

Some basic function for calculating summary statistics include **summary**, **mean()**, **median()**, **var()**, and **sd()**.

```{r}
summary(peyton)
mean(peyton)
sd(peyton)
```

R allows you to install additional packages (collections of functions) that aren't offered in the base version of R. To install a package, use **install.packages(<name_of_package>)** and to load a package, use **library(<name_of_package>)**.

One package that we will use frequently is **tidyverse**. This package includes several other packages and functions such as **ggplot** (plotting function), **dplyr** (data manipulation package), and **stringr** (string manipulation package).

```{r,message=F,include=F}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)
```

```{r,message=F,eval=F}
install.packages("tidyverse")
library("tidyverse")
```

You will also need to know how to load datasets from files. For this class, we will typically provide data files is .csv format.

Here is how to load a file:
```{r load, message=F}
# load readr package and load example dataset
library(readr)
NFL_2021_Team_Passing <- read_csv("data/NFL_2021_Team_Passing.csv")

# we can look at the header (first few entries) using "head()"
head(NFL_2021_Team_Passing)
```

-----

## Descriptive Statistics

### Definitions

::: {.definition}
A ***population*** is a well-defined complete collection of objects.
:::

::: {.definition}
A ***sample*** is a subset of the population.
:::

::: {.example}
Suppose we are interested in studying Peyton's Manning's season passing yards totals. How could you define the population and what is one possible sample?
::: 

\
\
\
\
\



::: {.definition}
***Quantitative data*** is numeric data or numbers. It can be broken into two further categories: discrete and continuous data. 
:::

::: {.definition}
***Discrete data*** is quantitative data with a finite or countably infinite number of values. 
:::

::: {.definition}
***Continuous data*** is quantitative data with an uncountably infinite number of values or data taken from an interval.
:::

::: {.example}
What are possible discrete and continuous data associated with Peyton Manning?
:::

\
\
\
\
\

::: {.definition}
***Qualitative data*** refers to names, categories, or descriptions. It can also be broken down into
two further categories, nominal data and ordinal data.
:::

::: {.definition}
***Nominal data*** is qualitative data with no natural ordering.
:::

::: {.definition}
***Ordinal data*** is qualitative data with a natural ordering.
:::

::: {.example}
What are possible nominal and ordinal data associated with Peyton Manning?
:::

\
\
\
\
\

-----


### Descriptive Statistics

While we will learn about some descriptive statistics that are unique to specific sports, there are some descriptive statistics that are frequently used in many applications.

#### Descriptive Statistics for Quantitative Data

There are different descriptive statistics depending on the type of data you are analyzing. We will begin by looking at descriptive statistics for quantitative data.

To begin, let $x_1, x_2, \ldots, x_n$ represent a numerical dataset with a sample of size $n$, where $x_i$ is the $i^\text{th}$ value in the dataset.

::: {.definition}
The ***sum*** of the data values is given by: $\sum_{i=1}^n x_i = x_1 + x_2 + \ldots + x_n$
:::

::: {.definition}
The ***sample mean*** (or sample average), $\bar{x}$, of the numerical dataset is given by $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i$
:::

::: {.definition}
The ***population mean*** (or population average), $\mu$, is the mean value for the entire population.
:::

The mean can be thought of as a measure of center or more generally, a measure of location.

::: {.example}
Recall that Peyton Manning's season passing yards total while with the Broncos were: 4659, 5477, 4727, 2249. Calculate the sample mean of these values.
:::

\
\
\
\
\

```{r mean}
# Calculate the sample of Peyton Manning's passing yards season totals with Colts
peyton.broncos <- c(4659, 5477, 4727, 2249)
mean(peyton.broncos)
```

In sports statistics, we often have to choose between using a descriptive statistic that summarizes a quantity versus a descriptive statistic that summarizes a rate. For instance, in basketball, we can compare two players based on how many points they score in a game (total quantity) or we can compare two players based on how many points per minute played (rate statistic). Many applications in sports analytics focus more on rate statistics rather than quantity statistics. Why?

We can measure the spread or variability of a dataset using *variance* and *standard devatiation*.

::: {.definition}
The ***sample variance***, $s^2$, of the numerical dataset is a measure of spread and is given by $s^2 = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2$
:::

::: {.definition}
The ***sample standard deviation***, $s$, of the numerical dataset is a measure of spread and is given by $s = \sqrt{s^2} = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2}$
:::

::: {.definition}
The ***population variance***, $\sigma^2$, is the variance for an entire population.
:::

::: {.definition}
The ***population standard deviation***, $\sigma$, is the standard deviation for an entire population.
:::

We often prefer to work with standard deviations as a measure of spread as opposed to variance because standard deviations are given in our original units.

```{r var}
# Calculate the variance and standard deviation of Peyton Manning's passing yards season totals with Broncos
var(peyton.broncos) # units: yards^2
sd(peyton.broncos) # units: yards
```

::: {.definition}
The **sample median**, $\tilde{x}$, of a numerical dataset is the middle value when the data are ordered from smallest to largest. In other words, let $x_1, x_2, \ldots, x_n$ be the (unordered) dataset and let $x_{(1)},x_{(2)}, \ldots, x_{(n)}$ be the same dataset but ordered from smallest to largest. If $n$ is odd, then $\tilde{x} = x_{(n+1)/2}$ and if $n$ is even, then $\tilde{x} = \frac{1}{2} \cdot \left[x_{\left(\frac{n}{2}\right)} + x_{\left(\frac{n+1}{2}\right)}\right]$.
:::


::: {.example}
Calculate the sample median of Peyton Manning's season passing yards total while with the Colts (3739, 4135, 4413, 4131, 4200, 4267, 4557, 3747, 4397, 4040, 4002, 4500, 4700). 
:::

\
\
\
\
\

Like sample mean, sample median is a measure of center. It gives you an idea of where the ``middle" of your dataset is. 

We can calculate sample mean and sample median in R as follows:

```{r median}
# Calculate the median of Peyton Manning's passing yards season totals with Broncos and Colts
peyton.colts <- c(3739, 4135, 4413, 4131, 4200, 4267, 4557, 3747, 4397, 4040, 4002, 4500, 4700)
median(peyton.broncos)
median(peyton.colts)
```

:::{.definition}
A ***percentile*** is a measure of relative standing. The $p^\text{th}$ percentile is the number where at least
p\% of the data values are less than or equal to this number.
:::

:::{.definition}
A ***quantile*** is a measure of relative standing and are the cut points for breaking a distribution of values into equal sized bins.
:::

:::{.definition}
A ***quartile*** is a measure of relative standing and are the cut points for breaking a distribution of values into four equal parts.
:::


```{r perc}
# Calculate the 10th and 90th percentile of Peyton Manning's passing yards season totals with Colts
quantile(peyton.colts,0.10)
quantile(peyton.colts,0.90)
quantile(peyton.colts,c(0.1,0.9))
```

**Special percentiles:** \
1. 25th percentile = 1st quartile = $Q_1$ \
2. 50th percentile = 2nd quartile = $Q_2$ = $\tilde{x}$ \
3. 75th percentile = 3rd quartile = $Q_3$ \

:::{.definition}
***Range*** is a measure of spread, measures the full width of a dataset, and is given by: $Range = Max - Min$.
:::

:::{.definition}
***Interquartile range*** is a measure of spread, measures the width of the middle 50\% of a dataset, and is given by: $IQR = Q_3 - Q_1$. 
:::

:::{.definition}
A ***five number summary*** describes the center, spread, and edges of a dataset and is given by: $(Min,Q_1,Q_2,Q_3,max)$.
:::

```{r iqr}
summary(peyton.colts)
quantile(peyton.colts,c(0,0.25,0.5,0.75,1))
```

#### Descriptive Statistics for Qualitative Data

In sports statistics, we also encounter qualitative (categorical) data which is names or labels which has its own descriptive statistics.


To begin, let $x_1, x_2, \ldots, x_n$ represent a categorical dataset with a sample of size $n$, where $x_i$ is the $i^\text{th}$ value in the dataset.

::: {.definition}
The ***proportion*** of sampled data that fall into a category is given by: $p = \frac{\#\text{ in category}}{\#\text{ total}}$
:::

''Proportion" and "Probability" are often used interchangeably. Both have a minimum value of 0 and a maximum value of 1.

::: {.definition}
The ***percentage*** of sampled data that fall into a category is given by: $P\% = 100 \cdot p = 100 \cdot \frac{\#\text{ in category}}{\#\text{ total}}$
:::

Percentages in this context can have a minimum value of 0\% and a maximum value of 100\%.

::: {.example}
In 2014, Peyton Manning started as quarterback for the Denver Broncos. The result of the Broncos' 16-game season was: \
Win, Win, Loss, Win, Win, Win, Win, Loss, Win, Loss, Win, Win, Win, Win, Loss, Win

Calculate the proportion and percentage of Broncos' winning games in 2014.
:::

\
\
\
\
\

```{r prop}
broncos2014 <- c("Win", "Win", "Loss", "Win", "Win", "Win", "Win", "Loss", "Win", "Loss", "Win", "Win", "Win", "Win", "Loss", "Win")
broncos.prop <- sum(broncos2014 == "Win")/length(broncos2014); broncos.prop
broncos.perc <- 100*broncos.prop; broncos.perc
```

We can also build a frequency table that summarizes the categories and their occurrences using **table()** in R. Note that **table()** works for quantitative and qualitative data.

```{r freq}
table(broncos2014)
```

## Visualizations

Conveying information visually is also an important part in providing a description of a dataset.

R provides some basic plotting functions such as **plot**, **hist**, and **barplot**. These plotting functions are simple and not always very clean looking.

In this class, we will use analogous plotting functions in **ggplot2** that are much improved plotting functions.

If you have already installed the **tidyverse** package, it should have also installed the **ggplot2** package.

```{r setup1}
# You have likely already installed the tidyverse package but if not, use the following command without the "#"
# install.packages("tidyverse")

# Load the tidyverse package (which includes ggplot2)
library(tidyverse)
```

Let's load the file ``NFL_2021_Team_Passing.csv" which contains [NFL Team Passing Statistics, 2021](https://www.pro-football-reference.com/years/2021/index.htm#passing)

```{r load2, message=F}
library(readr)
NFL_2021_Team_Passing <- read_csv("data/NFL_2021_Team_Passing.csv")
```

Histograms are one of the most common and basic ways to visualize a dataset's distribution of values. To make a histogram, you will use **ggplot** and **geom_histogram**.

:::{.example}
Create a histogram of the NFL Team Passing Yards in 2021.

```{r hist, message=F}
NFL_2021_Team_Passing %>% ggplot(aes(x=Yds)) + geom_histogram()
```

Notice how **%>%** is used to **pipe** the dataset into ggplot. This is using the pipe function from the **dplyr** package.

By default, **geom_histogram** uses 30 bins but this is customizable. Let's make the bins have a width of 200.

All good visualizations have good labels. Let's improve the axis labels and give the figure a title.

```{r hist2}
NFL_2021_Team_Passing %>% ggplot(aes(x=Yds)) + 
  geom_histogram(binwidth = 200) +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021")
```

We also have numerous options to change the appearance of plots when using **ggplot**. Let's change the bins color to *blue* and change the bin borders to *white*.

```{r hist3}
NFL_2021_Team_Passing %>% ggplot(aes(x=Yds)) + 
  geom_histogram(color = "white", fill = "blue", binwidth = 200) +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021")
```
:::

We can also create bar plots using ggplot using the **geom_bar** function.

:::{.example}
Create a bar plot with teams on the horizontal axis and passing touchdowns on the vertical axis.

```{r bar}
NFL_2021_Team_Passing %>% ggplot(aes(x=Tm,y=Yds)) +
  geom_bar(stat="identity")
```

The team labels are a complete mess. Let's fix this and make some adjustments to the axis labels and figure title.

```{r bar2}
NFL_2021_Team_Passing %>% ggplot(aes(x=Tm,y=Yds)) +
  geom_bar(stat="identity") +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

We can flip this graph if we like as well. Note that when we flip the graph, our labels get in reverse ordering, so this can be fixed using **fct_rev()** which is part of the **forcats** package.

```{r bar3}
NFL_2021_Team_Passing %>% 
  ggplot(aes(x=fct_rev(Tm),y=Yds)) +
  geom_bar(stat="identity") +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021") +
  coord_flip()
```

:::

Another common and useful visualization is a scatterplot which shows the relationship between two numeric variable. In ggplot, you use **geom_point()**.

:::{.example}
Create a scatterplot of Team Passing Yards and Team Passing Touchdowns from the NFL 2021 dataset.

```{r scatter1}
NFL_2021_Team_Passing %>%
  ggplot(aes(x=Yds,y=TD,label=Tm)) + 
  geom_point() +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021")
```

We may want to include team labels on this plot, however, it can get messy very quickly with a lot of points. 

```{r scatter2}
NFL_2021_Team_Passing %>%
  ggplot(aes(x=Yds,y=TD,label=Tm)) + 
  geom_point() +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021") +
  geom_text()
```

Many sports leagues have around 30 teams, so a clean scatterplot with labels can be tricky to make. Here are some options below.

```{r scatter3}
# install ggrepel package
library(ggrepel)
NFL_2021_Team_Passing %>%
  ggplot(aes(x=Yds,y=TD,label=Tm)) + 
  geom_point() +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021") +
  geom_text_repel()
```

```{r scatter4}
NFL_2021_Team_Passing$Abbr <- c("TB","LAC","DAL","KC","LAR","LV","CIN","GB","BUF","AZ","MN","SF",
                                "BAL","NE","PIT","ATL","MIA","DET","DEN","NYJ","WAS","JAC","SEA",
                                "TEN","PHI","IND","CLE","HOU","CAR","CHI","NYG","NO")
NFL_2021_Team_Passing %>%
  ggplot(aes(x=Yds,y=TD,label=Abbr)) + 
  geom_point() +
  labs(x="Team Passing Yards",y="Team Passing Touchdowns",title="NFL Team Passing Yards, 2021") +
  geom_text_repel(box.padding = 0.3) 
```


:::

-----

## Baseball

## Football

## Basketball

## Soccer

## Volleyball

## Hockey

For this example, we'll use a set of NHL data from moneypuck.com. First, let's load the data into `R` and open the data frame.

```{r hockey_ggplot1, message=F, warning=F}
nhl_2022_data <- read_csv("https://moneypuck.com/moneypuck/playerData/seasonSummary/2021/regular/teams.csv")

head(nhl_2022_data)
```

We can create nice looking tables using the ``kableExtra'' package. Let's look at the first eight rows and a samll selection of columns of the data frame and format the table output using a kable table.

```{r hockey_ggplot1a}
library("kableExtra")

nhl_2022_data[1:8, c(3,6:9)] %>% kbl() %>% kable_styling()
```

This dataset includes a *lot* of covariates. It also splits these data by different game situations: even-strength (5 on 5), power play (5 on 4), etc. Let's subset the data to include all game situations.

Use the `nrow` command to check the number of columns in the new data frame. Check: Is it the same as the number of teams in the league for the 2021-2022 season?

```{r hockey_ggplot2}
nhl_data_all <- filter(nhl_2022_data, situation == "all")

nrow(nhl_data_all)
```

The dataset includes an Expected Goals statistic for each team in the `xGoalsFor` column. Let's plot this quantity against the team's actual number of goals scored; this is given by the `goalsFor` column.

(Remember to always have a good title and axis labels!)

```{r hockey_ggplot3}
ggplot(data=nhl_data_all, aes(x=xGoalsFor, y=goalsFor)) + labs(x="Expected Goals Scored", y="Actual Goals Scored", title="NHL Teams: Expected vs. Actual Goals, 2021-22") + geom_point()
```

As expected, there is a general positive correlation between expected and actual goals ($r \approx 0.8$). However, there is some variability - for example, the Kings only scored 7 more actual goals than the Ducks, despite having 56.6 more expected goals.

Let's add a line to the graph using the `geom_abline` function corresponding to the line $y=x$, the line on which data points would fall if expected goals were equal to actual goals. We can also customize the line's color and type.

```{r hockey_ggplot4}
ggplot(data=nhl_data_all, aes(x=xGoalsFor, y=goalsFor)) + labs(x="Expected Goals Scored", y="Actual Goals Scored",title="NHL Teams: Expected vs. Actual Goals, 2021-22") + geom_point() + geom_abline(intercept=0, slope=1, color="red", linetype="dashed")
```

*Note: A slope of 0 and an intercept of 1 are actually the default parameters for the function.*

Q: What does it mean for a team's data point to fall below this line? Above it?

A: If the data point is below the line, it means the expected goals were greater than the actual goals; if the data point is above the line, it means the actual goals were greater than the expected goals. 

Q: Do you think that a team's expected goals would be more likely to be closer to its actual goals for a ten-game stretch, an entire season, or five consecutive seasons? Why?

<!-- Question could be reworded/clarified to mean "closer" in a relative (ratio) sense, rather than an absolute sense. -->

A: We would expect that as sample size increases, the result would become closer to expectation. So, actual goals would be most likely closer to expected goals over a span of five seasons.

<!--- This is basically conceptualizing the LLN, but that would rely on this calculation of expected goals being an unbiased estimator of actual goals. -->

