---
output:
  pdf_document: default
  html_document: default
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(formatR)
library(tidyverse)
library(kableExtra)

# Set so that long lines in R will be wrapped:
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=FALSE)
knitr::opts_chunk$set(cache=TRUE)

# kable table global setup
kt <- function(data) {
  knitr::kable(data, digits=3,linesep='',booktabs=TRUE) %>% kable_styling(bootstrap_options='striped', latex_options='HOLD_position', full_width = F, position = "center")
}
```


# Odds and Bets

**Sports Betting in USA**

In 2018, the United States Supreme Court overturned a 1992 federal law that banned commercial sports betting in most states. For more about this ruling, see this New York Times article: https://www.nytimes.com/2018/05/14/us/politics/supreme-court-sports-betting-new-jersey.html

In Colorado, you must be at least 21 years old to gamble including betting on sports games.

**Gambling Addiction**

Gambling addiction, compulsive gambling, or gambling disorder is a serious impulse-control disorder. Here is a link to the Mayo Clinic's discussion of compulsive gambling: https://www.mayoclinic.org/diseases-conditions/compulsive-gambling/symptoms-causes/syc-20355178

Gambling is ***not*** encouraged and can lead to numerous problems. If you choose to gamble on sports, don't bet beyond your means and seek help if you worry that you may be suffering from gambling addiction.

\newpage

## Odds

In the previous chapter, we quantified uncertainty using probabilities (numbers between 0 and 1) and percentages (numbers between 0 and 100). We can also quantify uncertainty using **odds**.

:::{.definition}
The ***odds*** (in favor) of an outcome $A$ is the probability of $A$ divided by the probability of $A^C$.

\[Odds = \frac{p}{1-p}\]
:::

**Note:** We may also be given the **odds against** a particular outcome. In this case, we have $Odds Against = \frac{1-p}{p}$.\

**Some Common Odds:**

```{r, echo=F}
a1 <- c(0:5,10,100)
a2 <- c(rep(1,8))
b1 <- c(rep(1,8))
b2 <- c(0:5,10,100)
p1 <- a1/(a1+b1)
p2 <- a2/(a2+b2)
q1 <- b1/(a1+b1)
q2 <- b2/(a2+b2)
odds1 <- c("0:1","1:1","2:1","3:1","4:1","5:1","10:1","100:1")
odds2 <- c("1:0","1:1","1:2","1:3","1:4","1:5","1:10","1:100")
odds_df <- data.frame(Odds = odds1, p = p1, q = q1, Odds = odds2, p = p2, q = q2)
names(odds_df) = c("Odds","p","q","Odds","p","q")
odds_df %>% kt()
```

In the following table, the odds of an outcome along with the probability of the outcome, *p*, and the probability of the outcome's complement, *q=1-p* are given.

\newpage

:::{.example}
For the following examples, convert between probability and odds.
:::

(a) Suppose that the probability that the Cubs win the 2025 World Series is 0.03. What are the odds in favor? \

\vfill

(b) Suppose that the odds that the Rockies win the World Series in 2025 is 100:1. What is the probability in favor? \

\vfill

(c) Suppose there is a 1\% chance that CSU wins a national championship in any sport in the next decade. What are the probability and the odds against CSU winning a national championship in the next decade?\

\vfill

:::{.example}
Suppose that there are three finalists in an Olympic competition. It is estimated that Athlete A has a 50\% chance of winning, Athlete B has a 40\% chance of winning, and Player C has a 10\% chance of winning. Calculate the odds against each of the athletes winning the competition.
:::

\vfill

\newpage

## Gambling Odds

Here's a helpful resource on calculating gambling odds: \
https://www.actionnetwork.com/education/decimal-odds

Gambling odds are a bit different that odds in a probability context. Gambling odds tell the amount that the bookmaker will pay out for a winning bet. For example, if a bookmaker is offering "10:1" that the Rockies win the next World Series, the bookmaker will pay out 10x the original wager plus the original wager if the Rockies win the World Series and the bookmaker will keep the original bet if the Rockies fail to win the World Series.

:::{.definition}
The ***implied probability***, $\tilde{p}$, of a wager is the probability that corresponds to the odds of the wager.

\[\tilde{p} = \frac{Odds}{Odds+1}\]
:::

:::{.definition}
***Fractional odds*** are common in horse racing and quote the net total that will be paid out to the bettor, should he or she win, relative to the stake. The numerator and denominator of fractional odds are always positive integers. For example, a \$100 wager on a "5:1" bet would result in a payout of $5 \cdot \$100 + \$100 = \$600$ if the wager is correct and a payout of \$0 if the wager is incorrect.
:::

:::{.definition}
***Decimal Odds*** represent the multiplier of a winning bet and is calculated by taking the inverse of the implied probability. For instance, if you bet \$100 with decimal odds of 1.8, your payout is \$180 (\$100 wager + \$80 winnings).

\[Decimal \, Odds = [\tilde{p}]^{-1}\]
:::

\newpage

## Types of Bets

:::{.definition}
A ***moneyline bet*** are common bets in American sports. When the moneyline is positive, the figure tells what the winning payout would be on a \$100 bet. When the moneyline is negative, the figure tells what wager is required for a winning payout of \$100.
:::

:::{.example}
For the following scenarios, assume a wager of \$100. Calculate the payout for a winning bet.
:::

(a) Fractional odds of 4/1 \

\vfill

(b) Moneyline +400 \

\vfill

(c) Moneyline -300 \

\vfill

(d) Decimal Odds of 1.5 \

\vfill

\newpage


:::{.definition}
A *moneyline bet* refers to the odds of a straight-up outcome on a game without consideration of a point spread. Typically, favorites will have negative moneyline odds and underdogs will have a positive moneyline odds.

If ML is negative:
\[\tilde{p} = \frac{-ML_{NEG}}{-ML_{NEG}+100}\]

If ML is positive:
\[\tilde{p} = \frac{100}{ML_{POS}+100}\]
:::

:::{.example}
Calculate the implied probabilities for the following examples.
:::

(a) Fractional odds of 4/1 \
\
\
\
\vfill

(b) Moneyline -125 \
\
\
\
\vfill

(c) Moneyline +125 \
\
\
\
\vfill

(d) Decimal Odds of 1.5 \
\
\
\
\vfill

\newpage

:::{.example}
Suppose that for an upcoming NFL game, a moneyline wager on the Broncos is -105 and a moneyline wager on the Raiders is -120.
:::

(a) Calculate the implied probabilities of the two possible outcomes. (Note that sometime ties are an optional wager. Other times, ties will result in a push, that is, no winner.) \
\
\
\
\vfill

(b) What is the sum of the implied probabilities? \
\
\
\
\vfill

(c) Does the result from (b) violate the axioms of probability? If so, why? \
\
\
\
\vfill


**Note:** The goal of bookmaking is to make money, so there will be almost certainly be a house advantage. This means that the sum of the implied probabilities will be greater than 1.

:::{.definition}
The ***house advantage*** (or ***hold percentage***) is the percentage of money that sportsbooks keep for every dollar earned.

\[House \, Advantage = 100 \cdot \sum_{i=1}^n \tilde{p}_i - 100\]
:::

\newpage

:::{.example}
Suppose that a bookmaker is offering a moneyline wager on the Broncos at -105 and a moneyline wager on the Raiders at -120. Calculate the bookmakers expected winnings based on the following amounts of total bets.
:::

(a) \$500 wagered on the Broncos, \$500 wagered on the Raiders \
\
\
\
\vfill

(b) \$250 wagered on the Broncos, \$750 wagered on the Raiders \
\
\
\
\vfill

(c) \$750 wagered on the Broncos, \$250 wagered on the Raiders \
\
\
\
\vfill

(d) \$0 wagered on the Broncos, \$1000 wagered on the Raiders \
\
\
\
\vfill

(e) \$1000 wagered on the Broncos, \$0 wagered on the Raiders \
\
\
\
\vfill

(f) What is the house advantage? \
\
\
\
\vfill

\newpage

:::{.example}
A ***point spread bet*** is a bet based on the projected margin of victory that can result in a win, loss, or push.

For example, if a sportsbook offers Rockies -1.5 against the Cubs and you bet on the Rockies, if the Rockies win by 2 or more runs, you win and if the Rockies win by 1 run or lose, then you lose.
:::

:::{.example}
Suppose a sportsbook offers Broncos +14 against the Raiders at -110 (they also offer Raiders -14 at -110) and you have \$20 to wager. Calculate the payout for the following scenarios.
:::

(a) You bet \$20 on the Broncos and the final score is Broncos 21 Raiders 20. What is your payout? \
\
\
\
\vfill

(b) You bet \$20 on the Raiders and the final score is Broncos 21 Raiders 20. What is your payout? \
\
\
\
\vfill

(c) You bet \$10 on the Broncos and \$10 on the Raiders and the final score is Broncos 21 Raiders 20. What is your payout? \
\
\
\
\vfill

:::{.example}
A ***parlay bet*** is a combination of multiple bets where the winnings from a winning bet are placed on other bets. All bets must be win for the parlay bet to pay out.
:::

\newpage

:::{.example}
Suppose that CSU is a three point favorite (-3) against CU in a women's basketball game and the over/under on total number of points is 97. Assume that the price of bets is -110.
:::

(a) Suppose you place two bets, CSU -3 and Over 97, and the outcome of the game is CSU 60 CU 40. What is your payout? \
\
\
\
\vfill

(b) Suppose you place a parlay bet on CSU -3 and Over 97 and the outcome of the game is CSU 60 CU 40. What is your payout? \
\
\
\
\vfill

(c) Suppose you place two bets, CSU -3 and Over 97, and the outcome of the game is CSU 44 CU 40. What is your payout? \
\
\
\
\vfill

(d) Suppose you place a parlay bet on CSU -3 and Over 97 and the outcome of the game is CSU 44 CU 40. What is your payout? \
\
\
\
\vfill




*Reference:* \
https://www.wikihow.com/Calculate-Odds

**Class Reading:** Chapters 44 in the *Mathletics* by Wayne Winston