```{r, echo=F, message=F,warning=F}
library(tidyverse)
library(kableExtra)
knitr::opts_chunk$set(tidy=FALSE)

# kable table global setup
kt <- function(data) {
  knitr::kable(data, digits=3, align=c('l','c','c','c','c','c','c','c','c')) %>% kable_styling(bootstrap_options='striped', latex_options='HOLD_position', full_width = F, position = "center")
}
```

# Correlation

## Pearson's Correlation Coefficient

### Partial Correlation and Confounding Variables

(AM, page 131, pitchers with more innings)

### Properties of the Linear Correlation Coefficient

Pearson's correlation coefficient is unaffected by changes in location (adding a constant) or scale (multiplying by a constant) of data. To illustrate this concept, let's consider the correlation between the number of points scored by the winning team of the Super Bowl and 1) the number of the Super Bowl {1, 2, 3, ... , 56} and 2) the year of the Super Bowl {1967, 1968, 1969, ... , 2022}.

```{r correlation properties, include=FALSE}
Super_Bowl_Number <- c(1:56)
Super_Bowl_Year <- c(1967:2022)
Winning_Team_Points <- c(35,33,16,23,16,24,14,24,16,21,32,27,25,31,27,26,27,38,38,46,39,42,20,55,20,37,52,30,49,27,35,31,34,23,34,20,48,32,24,21,29,17,27,31,31,21,34,43,28,24,34,41,13,31,31,23)

Super_Bowl_Data <- data.frame(Super_Bowl_Number, Super_Bowl_Year, Winning_Team_Points)
```

```{r correlation properties 2, echo=FALSE}
ggplot(Super_Bowl_Data, aes(x=Super_Bowl_Number, y=Winning_Team_Points)) + 
    geom_point() +
    geom_line(color="red") +
    ggtitle("Super Bowl Number and Points Scored by Winning Team")

ggplot(Super_Bowl_Data, aes(x=Super_Bowl_Year, y=Winning_Team_Points)) + 
    geom_point() +
    geom_line(color="red") +
    ggtitle("Super Bowl Number and Points Scored by Winning Team")
```

```{r correlation properties 3}
cor(Super_Bowl_Number, Winning_Team_Points)
cor(Super_Bowl_Year, Winning_Team_Points)
```

The only difference between the two plots is that the x-axis is shifted. Since the change is essentially a linear (location) transformation, the correlations are identical.

Note: The correlations were equal only because there has been a one-to-one relationship between Super Bowl number and year since 1967. Major League Baseball has had years in which no World Series was held (1904 - boycott, 1994 - players' strike); thus, the same example applied to the MLB would produce slightly different correlation coefficients.

Q: The correlation coefficient is positive. What does this mean in terms of the trend of number of points scored by Super Bowl winners over time?

A: Winners of more recent Super Bowls are more likely to have scored more points than winners of older Super Bowls.

In a similar way to the example above, the correlation between a basketball player's career field goal attempts and three-pointers made and the correlation between the same player's career FGA and points scored from three-point field goals are exactly the same, due to the scale invariance property of correlation.

<!-- Note to Aaron: This could be a good homework exercise. Find data or simulate career FGA and 3PM for basketball players, then show that the two correlations described above are equal. -->

## Rank Correlation

<!-- Used for nonlinear relationships -->

(Example using team data)

## Autocorrelation

(AM, page 134, NFL team records, pre- and post- salary cap)

### Hot Hand

(Scorecasting, page 215)

### Streakiness in Sports

(Mathletics, chapter 11, page 105)

## Association of Categorical Variables

(AM, page 143, Yule's Q)

(AM, page 139, winning at halftime vs winning game)

(AM, page 144, Brady TD passes vs sacks)

(AM, page 145, Nadal clay courts)
