# Aaron's stuff

## Notes for Chapter 2 (Probability)

__Axioms of Probability:__  

1. $P(A) \geq 0$

2. $P(\Omega)=1$

3. If $A_1, A_2, \ldots, A_n$ are disjoint events, then $P(\cup_{i=1}^n A_i) = \sum_{i=1}^n P(A_i)$

::: {.theorem name="Bayes theorem"}
Let $A$ and $B$ be events in $\Omega$ such that $P(B)>0$. Then we have the following:
$$
P(A|B) = \frac{P(B|A)P(A)}{P(B)}
$$
:::

## Suggested Readings

### Moneyball

Moneyball, Chapter 2, How to Find a Ballplayer [@lewis2004moneyball]

Near the end of the chapter (page 40), Michael Lewis give a list of players the Oakland Athletics hoped to draft. How did these players turn out? Find the WAR for each of the players in their pre-free agency years and compare it against the Rockies draft picks in the same rounds from the same draft.

### Future Value

Future Value, Chapter 7, How to Scout [@longenhagen2020future]

If a player receives a running grade of 40, approximately what proportion of MLB players have a lower have a lower running grade?

For a given tool, about 95\% of all player grades fall between what two bounds? (Consider the middle 95\% of the distribution of grades.)

## Notes for Chapter 4 (Simulation)

### Baseball Simulation Example

```{r setup,warning=F,message=F}
library(tidyverse)
```

This is a baseball example for chapter 4.

```{r baseball-sim, message=F}
set.seed(2022)
n.sims <- 1000
hits <- rep(0,n.sims)
avg <- 0.300
atbats.mean <- 450
atbats.sd <- 100
sim.atbats <- round(rnorm(n.sims,atbats.mean,atbats.sd))

for(i in 1:n.sims){
  sim.hits <- rbinom(1,sim.atbats[i],avg)  
  hits[i] = sim.hits
}
hits.df <- data.frame(Hits=hits) 
hits.df %>% ggplot(aes(x=Hits)) + geom_histogram()
``` 

<!-- usethis::use_git() -->
<!-- git push --set-upstream origin master -->
