# Other Sports

## Simulation

### Shaq FT simulation

:::{.example}
In 1997-1998 with the Los Angeles Lakers, Shaq attempted an average of 11.35 free throws per game with a standard deviation of 4.04. While with the Lakers, Shaq played in an average of 63.6 games per year with a standard deviation of 10.6. Create a simulation to model the season total number of free throw attempts that Shaq would have while with the Lakers. 

Note: his actual season totals of free throws attempted while with the Lakers were: 479, 681, 498, 824, 972, 712, 725, 676
:::

Let's model the number of games that Shaq played in as a Binomial random variable. There are 82 regular sesaon games, so let $n=82$. Shaq played in an average of 63.6 games, so let $p=\frac{64.25}{82}=0.784$. Shaq played in about 78% of the games during his career with the Lakers.

For the number of free throw attempts per game, we could model this as a Poisson random variable or a Negative Binomial random variable. As noted in the previous chapter, the variance of Shaq's FT attempts is a fair bit greater than the mean which means that it is overdispersed. Negative Binomial may be a more appropriate model than a Poisson. From last chapter, we found $\hat{r}=25.85$ and $\hat{p}=0.305$.

```{r}
set.seed(2020)
n.sims <- 10000
FT <- rep(0,n.sims)

# simulate the number of games played in a season, round to a whole number
games.sim <- rbinom(n = n.sims,size = 82,prob = 0.784)

# number of games can't exceed 82 games (regular season total)
games.sim[games.sim>82] = 82

for(i in 1:n.sims){
  # simulate the season total FT attempts in each simulation
  temp <- rnbinom(n = games.sim[i],size = 25.85,prob = 1-0.305)
  FT[i] <- sum(temp)
}

# Simulated mean and SD of season totals of free throw attempts
c(mean(FT),sd(FT))

# Actual mean and SD of season totals of free throw attempts
FT.actual <- c(479, 681, 498, 824, 972, 712, 725, 676)
FT.actual.mean <- mean(FT.actual)
FT.actual.var <- var(FT.actual) * 7/8 # population variance
FT.actual.sd <- sqrt(FT.actual.var)
c(FT.actual.mean,FT.actual.sd)

FT %>% as.data.frame() %>% 
  ggplot(aes(x=FT)) + 
  geom_histogram(bins=30,color = "yellow", fill = "purple") +
  ggtitle("Simulated Season FT Attempt Totals by Shaq with Lakers")
```

Notice that the mean of our simulation is somewhat close to Shaq's true season average number of free throw attempts but the variance of the simulation is far too low. 

We can also simulate data using resampling. In this case, rather than simulating random variables according to a distribution, we can use our actual data as a sampling distribution.

:::{.example}
Using resampling, simulate the number of free throws Shaq would attempt while with the Lakers. Compare the mean and variance of the simulation to Shaq's actual statistics.
:::

```{r, message=F}
set.seed(2020)
n.sims <- 10000
FT <- rep(0,n.sims)
shaq.games <- c(51,60,49,79,74,67,67,67)
shaq.FTA <- read_csv("data/shaqFT.csv", col_names = FALSE)
shaq.FTA <- shaq.FTA$X1

# sample (with replacement) from Shaq's FTA game totals
games.sim <- sample(x = shaq.games,size = n.sims,replace = T)

for(i in 1:n.sims){
  # sample (with replacement) from Shaq's FTA game totals
  temp <- sample(x = shaq.FTA,size = games.sim[i],replace = T)
  FT[i] <- sum(temp)
}

# Simulated mean and SD of season totals of free throw attempts
c(mean(FT),sd(FT))

# Actual mean and SD of season totals of free throw attempts
FT.actual <- c(479, 681, 498, 824, 972, 712, 725, 676)
FT.actual.mean <- mean(FT.actual)
FT.actual.var <- var(FT.actual) * 7/8 # population variance
FT.actual.sd <- sqrt(FT.actual.var)
c(FT.actual.mean,FT.actual.sd)

FT %>% as.data.frame() %>% 
  ggplot(aes(x=FT)) + 
  geom_histogram(bins=30,color = "yellow", fill = "purple") +
  ggtitle("Simulated Season FT Attempt Totals by Shaq with Lakers")
```

This simulation is biased low on the variance but is better than the earlier simulation. 

One reason that you may want to do complicated simulations like the example above is to make predictions for a player's future seasons.

## Probability

### Sets and Conditional Probability

100 sports fans in Colorado were polled and it was found that 64 had attended either a Denver Nuggets or Colorado Avalanche game at Ball Arena (formerly Pepsi Center). 34 people had seen only a Nuggets game, while 17 had seen both a Nuggets and an Avalanche game.

Q: How many people saw an Avalanche game but not a Nuggets game?

A: 64 - 34 - 17 = 13

Q: What is the probability that a randomly selected person in the poll had been to a Nuggets game?

A: (34 + 17) / 100 = .51

Q: What is the probability that a randomly selected person that had been to a game at Ball Arena had been to a Nuggets game?

A: (34 + 17) / 64 = .797

Q: What is the probability that a randomly selected person had been to a Nuggets game given they had been to an Avalanche game?

A: 17 / (17 + 13) = .567


### Binomials and Multinomials

Suppose we are curious about probabilities regarding the results of a soccer team’s next five games.

Wait!!! A soccer game has three possible outcomes (win, lose, draw)! We can’t use the binomial distribution, since it limits us to two possible outcomes!

It depends. If we are interested in the probability that a soccer team wins 2 of their next 5 games, we can use the binomial distribution. We can create the following partition of the sample space of outcomes: $(Win)$ and $(Win^C)$, where the second set includes both losing and drawing.

Then, the formula would be represented as:

$\binom{5}{2}\ P(Win)^2\  P(Win^C)^{(5-2)}$

If we are interested in the probability of the team winning two of the next five games, drawing two, and losing one, we cannot use the binomial theorem. That involves three outcomes, and would be represented as a multinomial. <!--- assuming regularity conditions such as independence. -->



### Expectation - Baseball

The expectation of a discrete random variable is a weighted average. The "weights" are the probabilities of the possible values of the variable.

Consider the following table, which shows the number of career hits by type for the all-time Major League Baseball leader in total bases, Hank Aaron.

<!-- https://www.baseball-reference.com/players/a/aaronha01.shtml -->

```{r expectation baseball, echo=F}
Hit_type <- c("Single", "Double", "Triple", "Home Run")
Number_bases <- c(1:4)
Hit_Frequency <- c(2294, 624, 98, 755)
Hit_Proportion <- round(Hit_Frequency / sum(Hit_Frequency), 8)

Hank_Aaron_Hits <- data.frame(Hit_type, Number_bases, Hit_Frequency, Hit_Proportion)

Hank_Aaron_Hits %>%
kbl() %>%
kable_styling()
```

The expected number of bases for a Hank Aaron hit is the sum of the number of bases attained for each hit multiplied by the relative frequency of the occurrence of that type of hit.

$1 \cdot \frac{2294}{3771} + 2 \cdot \frac{624}{3771} + 3 \cdot \frac{98}{3771} + 4 \cdot \frac{755}{3771} = 1.18181$

This is the same process that is occurring whenever we calculate the expectation of any discrete random variable. Recall the formula for expectation is $E[X] = \sum_{x \in \Omega}\ x \cdot p(x)$. Each value in the sample space is "adjusted" by the probability of that value, then the sum of all values in $\Omega$ is taken to arrive at the weighted average, or expected value, of the random variable.

### Basketball Scenario

You are the coach of a basketball team that is down two points with one second remaining in the fourth quarter. During a timeout, you are considering the best play to call for your team. The first option is a three-point shot attempt, which you estimate has a 30% chance of succeeding. The second option is a two-point shot attempt, which has a 50% chance of making the field goal, a 30% chance of missing it and ending the game, and a 20% chance the shooter will miss but be fouled, in which case the shooter's free throw success will follow a $Bin(2, .8)$ random variable. Finally, you estimate that your team's probability of winning the game in overtime is .45.

Assume the above situations are exhaustive (i.e., the other team will not get another possession, no fouls will be called before the ball is put in play, lightning will not hit the arena and postpone the game, etc.). Which of the two plays should you call to maximize the win probability for your team?

A: The probability of winning the game with the three-point shot attempt is .3. If the two-point shot attempt is called for, there is a .5 probability of making the field goal and a (.2)(.8)(.8) = .128 probability that the foul is called and both free throws are made. Thus, the total probability of scoring two points and sending the game to overtime is .628. Then, the probability of winning the game in OT after tying it in regulation is (.628)(.45) = .2828. This is less than .3, so shooting the three-pointer is the option that maximizes the win probability, given these situational probabilities.

Q: What is the minimum estimated overtime win probability to make calling for the two-point play the better option?

A: $P(score\ 2\ points\ in\ regulation) \cdot P(win\ in\ OT) > P(win\ in\ regulation)$\
$.628 \cdot P(win\ in\ OT) > .3$\
$P(win\ in\ OT) > .478$

### Multiple Probability Distributions - Basketball

Suppose the number of points scored by a basketball player follows a Poisson(12) random variable, the number of rebounds by a Poisson(7) distribution, and assists by a Discrete Uniform(2, 11), independently of each other.

Q: What is the probability that this player records a points, rebounds, assists triple-double in a game?

A: $P(Triple\ Double) = P(Points \geq 10\ \cap\ Rebounds \geq 10\ \cap\ Assists \geq 10)$

```{r basketball probability 1}
ppois(9, lambda = 12, lower.tail=F)
```

$P(Points \geq 10) = P(Poisson(12) \geq 10) \approx .758$

```{r basketball probability 2}
ppois(9, lambda = 7, lower.tail=F)
```

$P(Rebounds \geq 10) = P(Poisson(7) \geq 10) \approx .170$

$P(Assists \geq 10) = P(Discrete\ Uniform(2, 11) \geq 10) = .2$

Since the events are independent, we can multiply their probabilities. The probability of the player scoring the triple-double is $(.758)(.170)(.2) = .0257$.

Q: Your friend offers you 4 to 1 that the player will not record a triple-double in their next 10 games. With the knowledge that the athlete's performance in a game is unaffected by performances in previous games, would you take the bet?

A: $P(no\ triple\ double) = 1 - .0257 = .9743$, so $P(no\ triple\ double\ in\ next\ 10\ games) = (.9743)^{10} = .771$

The odds of no triple-double are $\frac{.771}{1-.771} = 3.37$, so the bet of no triple-double at 4 to 1 odds is favorable.



*answers may vary for following questions*

Q: What differences do you notice between the actual results and the expected values based on the Poisson random variable?

A: There were fewer games in which the Rapids scored 4 or more goals than the model would indicate, yet the Rapids were shut out less often than the model would indicate.

Q: Even if the true population distribution of 2021 Rapids goals was truly a Poisson(1.5) random variable, why might the actual distribution of their goals differ from the probability mass function?

A: 34 is a relatively small sample size; random variables may not coincide with their expected values for finite sample sizes.

Q: What are the advantages of using the Poisson distribution to model Major League soccer goals? What are the disadvantages?

A: Poisson random variables can take on the natural numbers (including zero), which aligns with the number of goals that can be scored in a match. One disadvantage is that it is possible for a Poisson to take on values that are not realistic for the situation, such as double-digit integers or higher. Only one game in MLS history has had a team score more than seven goals in a game. However, when $\lambda$ is small (such as 1.5), these extreme values are relatively unlikely.


### Law of Total Probability - Baseball

You work in the front office of a professional baseball club and have just learned that a certain prospect hits .200 against left-handed pitchers and .400 against right-handed pitchers (their overall batting average is unknown). The general manager of the team overhears you talking about the .400 statistic of the player and becomes very exited that they have the chance to draft a .400 hitter. What would you say to caution the GM that the player might not be a remarkable hitter?

A: We don't know the proportion of the player's at-bats that came against left-handed pitchers versus right-handed pitchers. If we want to know the player's batting average unconditional on the type of pitcher they are facing, we have to adjust $P(hit\ |\ left-handed\ pitcher)$ by $P(left-handed\ pitcher)$ and $P(hit\ |\ right-handed\ pitcher)$ by $P(right-handed\ pitcher)$ before adding them to determine $P(hit)$. For example, if 90% of the player's at-bats were against left-handed pitchers, then their overall batting average is a pedestrian .220.

 
*Other possible issues: low sample size of player's at-bats, the fact that pro pitchers will be harder to hit against than non-pros*

### Multinomial Distribution - Baseball

The following PMF displays the number of possible bases for a hit in baseball against the frequency of that type of hit for Hank Aaron, who is third all-time in Major League baseball hits with 3,771.

```{r multinomial, include=F}
Number_bases <- c(1:4)
Hit_Frequency <- c(2294, 624, 98, 755)
Hit_Proportion <- round(Hit_Frequency / sum(Hit_Frequency), 5)

Hank_Aaron_PMF <- data.frame(Number_bases, Hit_Proportion); Hank_Aaron_PMF
```

| Number of Bases, $x$  | 1 | 2 | 3 | 4 |
|:---------------------:|:-:|:-:|:-:|:-:|
| Probability, $P(X=x)$ | 0.61 | 0.16 | 0.03 | 0.20 |

Q: What is the probability that a randomly selected subset of 5 hits from Hank Aaron includes three singles, one double, and one home run?

<!-- note: assumes sampling with replacement. Without replacement would be a multivariate hypergeometric and would slightly increase the probability.  -->

A: From the Multinomial PMF, this is $P(X_1 = 3, X_2 = 1, X_3 = 0, X_4 = 1) = \frac{5!}{3!\ \cdot 1!\ \cdot 0!\ \cdot 1!}\ (.61)^3\cdot\ (.16)^1\cdot\ (.03)^0\cdot\ (.20)^1 \approx .145$

The probability can also be found in `R` using the `dmultinom` function.

```{r multinomial 2}
dmultinom(x = c(3, 1, 0, 1), prob = c(.61, .16, .03, .20))
```


### Bayes - injured baseball player

:::{.example}
A runner on first base with 2 out and nobody else on base will attempt to steal second base on the first pitch 70% of the time if he is fully healthy but only 10% of the time if he is playing through an injury. Assume that 80% of the player population is healthy. You see a randomly selected runner not attempt a steal in this situation. What is the probability that the runner is playing through an injury?
:::

From Bayes Theorem:  

Pr(Injury given No Steal) = Pr(No Steal given Injury)\*Pr(Injury)/P(No Steal).  

Pr(No Steal given Injury) = 1 - Pr(Steal given Injury) = 0.9.  

Pr(Injury) = 1- Pr(Healthy) = 0.2.  

Pr(No Steal) = Pr(No Steal given Injury)\*Pr(Injury)+Pr(No Steal given Healthy)\*Pr(Healthy).  

Pr(No Steal) = 0.9\*0.2+0.7\*0.8 = 0.74.  

Therefore Pr(Injury given No Steal) = 0.9\*0.2/0.74 = 0.243.  

