---
output:
  html_document: default
  pdf_document: default
---
# Levi's stuff

```{r, include=FALSE}
library("tidyverse")
library("kableExtra")

kt <- function(data) {
   knitr::kable(data, digits=3, align=c('l','c','c','c','c','c','c','c','c')) %>% kable_styling(bootstrap_options='striped', latex_options='HOLD_position', full_width = F, position = "center")
}
```


<!-- ## Rules of Hockey -->

<!-- [Link to YouTube video describing hockey rules](https://www.youtube.com/watch?v=nv2FUnHceqU) -->

<!-- ## Basic Hockey Statistics -->

<!-- * **Goals (G)**: If a team scores, the skater on the scoring team who last touched the puck is credited with a goal. -->

<!-- * **Assists (A)**: The players (up to two) on the scoring team who last touch the puck before the goalscorer are credited with assists, unless the opposing team has possession of the puck in between. -->

<!-- * **Points (PTS)**: Goals plus assists. -->
<!--   * Not to be confused with team points awarded in the regular season standings by the many hockey leagues, including the NHL (two points for a win, one point for an overtime/shootout loss, zero points for a regulation loss). -->

<!-- * **Shots On Goal (SOG)**: Shot attempts in which the puck has been shot directly on goal. Shot attempts which are blocked or miss the goal are not considered SOGs. A team's shots on goal should equal the opposing goaltender's saves plus the team's goals scored. -->

<!-- * **Goals Against Average (GAA)**: Of a goaltender, the number of goals allowed by that goaltender adjusted to a per-60 minute rate. -->

<!-- * **Penalty Minutes (PIM)**: The amount of penalty time an individual player is assigned for their infractions. PIM may be different than the amount of time the player actually spends in the penalty box. -->

<!-- *Reference: https://www.milehighhockey.com/pages/stats* -->


<!-- ## Advanced Hockey Statistics -->

<!-- * **CORSI**: CORSI only applies to 5 on 5 ("even-strength") situations. It is calculated as the difference between shot attempts on offense (shots on goal + blocked shots + missed shots) minus shot attempts allowed on defense. CORSI can also be expressed as a percentage, with percentages over 50% indicating that the player is on ice for more offensive shots than defensive shots. -->

<!-- * **Expected Goals (xG)**: Expected Goals statistics give each shot an estimated probability of scoring a goal based on factors such as shot location and game situation. xG cannot be less than 0 or greater than 1 for any particular shot, and different platforms may have different methods of calculating expected goals. -->

<!-- * **Fenwick/Unblocked Shot Attempts (USAT)**: Similar to CORSI, but omits blocked shots from the calculation. This statistic is used in many Expected Goals calculations. -->

<!-- Because the flow of a hockey game is usually quite different in situations other than the normal 5 on 5, such as a power play (5 on 4) or concurrent penalties (4 on 4), many hockey databases separate data by the type of game situation. We will see this below with a dataset from [MoneyPuck](www.moneypuck.com), but it is also present on [Natural Stat Trick](https://www.naturalstattrick.com/), [QuantHockey](www.quanthockey.com), and [hockey-reference](www.hockey-reference.com). -->

<!-- *References: https://www.nhl.com/lightning/news/hockey-analytics-101-understanding-advanced-stats-and-how-theyre-measured/c-735819, https://theathletic.com/121980/2017/10/09/an-advanced-stat-primer-understanding-basic-hockey-metrics/* -->


<!-- ## Hockey - Expected and Actual Goals -->

<!-- For this example, we'll use a set of NHL data from moneypuck.com. First, let's load the data into `R` and open the data frame. -->

<!-- ```{r hockey ggplot 1} -->
<!-- nhl_2022_data <- read.csv("https://moneypuck.com/moneypuck/playerData/seasonSummary/2021/regular/teams.csv") -->

<!-- View(nhl_2022_data) -->
<!-- ``` -->

<!-- This dataset includes a *lot* of covariates. It also splits these data by different game situations: even-strength (5 on 5), power play (5 on 4), etc. Let's subset the data to include all game situations. -->

<!-- Use the `nrow` command to examine the number of columns in the new data frame. Check: Is it the same as the number of teams in the league for the 2021-2022 season? -->

<!-- ```{r hockey ggplot 2} -->
<!-- nhl_data_all <- filter(nhl_2022_data, situation == "all") -->
<!-- View(nhl_data_all) -->

<!-- nrow(nhl_data_all) -->
<!-- ``` -->

<!-- The dataset includes an Expected Goals statistic for each team in the `xGoalsFor` column. Let's plot this quantity against the team's actual number of goals scored; this is given by the `goalsFor` column. -->

<!-- (Remember to always have a good title and axis labels!) -->

<!-- ```{r hockey ggplot 3} -->
<!-- ggplot(data=nhl_data_all, aes(x=xGoalsFor, y=goalsFor)) + labs(x="Expected Goals Scored", y="Actual Goals Scored", title="NHL Teams: Expected vs. Actual Goals, 2021-22") + geom_point() -->
<!-- ``` -->

<!-- As expected, there is a general positive correlation between expected and actual goals ($r \approx 0.8$). However, there is some variability - for example, the Kings only scored 7 more actual goals than the Ducks, despite having 56.6 more expected goals. -->

<!-- Let's add a line to the graph using the `geom_abline` function corresponding to the line $y=x$, the line on which data points would fall if expected goals were equal to actual goals. We can also customize the line's color and type. -->

<!-- ```{r hockey ggplot 4} -->
<!-- ggplot(data=nhl_data_all, aes(x=xGoalsFor, y=goalsFor)) + labs(x="Expected Goals Scored", y="Actual Goals Scored",title="NHL Teams: Expected vs. Actual Goals, 2021-22") + geom_point() + geom_abline(intercept=0, slope=1, color="red", linetype="dashed") -->
<!-- ``` -->

<!-- *Note: A slope of 0 and an intercept of 1 are actually the default parameters for the `geom_abline` function.* -->

<!-- Q: What does it mean for a team's data point to fall below this line? Above it? -->

<!-- A: If the data point is below the line, it means the expected goals were greater than the actual goals; if the data point is above the line, it means the actual goals were greater than the expected goals. -->

<!-- Q: Do you think that a team's expected goals would be more likely to be closer to its actual goals for a ten-game stretch, an entire season, or five consecutive seasons? Why? -->

<!-- <!-- Question could be reworded/clarified to mean "closer" in a relative (ratio) sense, rather than an absolute sense. --> -->

<!-- A: We would expect that as sample size increases, the result would become closer to expectation. So, actual goals would be most likely closer to expected goals over a span of five seasons. -->

<!-- <!--- This is basically conceptualizing the LLN, but that would rely on this calculation of expected goals being an unbiased estimator of actual goals. --> -->

<!-- ## Hockey - Goalie Data -->

<!-- For this next example, let's use goalie data from the 2021-2022 season from [Natural Stat Trick](https://www.naturalstattrick.com/). -->

<!-- ```{r goalie data 1} -->
<!-- goalie_data <- read.csv("C:/Users/Fam/Downloads/GoalieTotals_NaturalStatTrick.csv") -->
<!-- View(goalie_data) -->
<!-- ``` -->

<!-- The dataset includes 119 goalies, but many of them didn't play very much. We can subset the data to include only goaltenders that faced at least 500 shots. -->

<!-- Which player among qualified goalies had the best goals against average? On which team did he play, and what was his GAA? -->

<!-- Which goalie had the most playing time? What was his team, and how much time did he spend on the ice? -->

<!-- ```{r goalie data 2} -->
<!-- filtered_goalie_data <- filter(goalie_data, Shots.Against >= 500) -->

<!-- filtered_goalie_data %>% filter(GAA == min(GAA)) %>% select(Player, Team, GAA) -->
<!-- filtered_goalie_data %>% filter(TOI == max(TOI)) %>% select(Player, Team, TOI) -->
<!-- ``` -->

<!-- The following plot compares save percentage to the number of shots on goal faced for the qualified goalies. The dashed horizontal line is placed at the average shots on goal faced among qualified players, and the dashed vertical line is placed at the average save percentage among qualified players. -->

<!-- ```{r goalie data 3, echo=FALSE} -->
<!-- ggplot(data=filtered_goalie_data, aes(x=SV., y=Shots.Against)) + labs(x="Save Percentage", y="Shots On Goal Faced",title="NHL Goalies: Save Percentage vs. Volume of Shots, 2021-22") + geom_point() + geom_hline(yintercept=mean(filtered_goalie_data$Shots.Against), color="red", linetype="dashed") + geom_vline(xintercept=mean(filtered_goalie_data$SV.), color="red", linetype="dashed") -->
<!-- ``` -->

<!-- Q: Which quadrant of the graph represents goalies that faced a higher than average number of shots, but had a below-average save percentage? -->

<!-- A: The second (upper left) quadrant. -->

<!-- Q: Which quadrant represents goalies that had a high save percentage and faced a high volume of shots? -->

<!-- A: The first (upper right) quadrant. -->

<!-- ## Hockey - Correlation Plots -->

<!-- When analyzing sports data, there may be many circumstances where statisticians consider which of several variables are most highly correlated to an outcome variable of interest. In this case, it can be useful to use a correlation plot (also known as a correlation matrix or correlogram). Tidyverse and related packages provide many options for creating correlation plots. -->

<!-- Suppose a statistician has recently learned about some advanced hockey statistics and is interested in researching which stat has the highest correlation with goals scored. The statistician wants to compare team shots on goal, CORSI, and Fenwick to observe the association with goals scored for NHL teams. -->

<!-- The following plot uses the same 2021-2022 data from Moneypuck.com; it gives the pairwise scatterplots and correlation values for each of the variables, as well as smoothed plots of each individual variable along the diagonals. -->

<!-- ```{r hockey ggplot 5} -->
<!-- library("GGally") -->

<!-- goal_stats <- nhl_data_all %>% select(shotsOnGoalFor, corsiPercentage, fenwickPercentage, goalsFor) -->
<!-- ggpairs(goal_stats, title="Correlation plot of goals and relevant predictors, NHL 2021-22") -->
<!-- ``` -->

<!-- Q: Which of the variables has the strongest correlation with goals scored? -->

<!-- A: CORSI percentage, r = .702. -->

<!-- In the article ["An advanced stat primer: Understanding basic hockey metrics"](https://theathletic.com/121980/2017/10/09/an-advanced-stat-primer-understanding-basic-hockey-metrics/), Charlie O'Connor states, "Generally speaking, Corsi is more predictive of future goal differential than Fenwick... however, Fenwick forms the basis for the most widely-used Expected Goals models." Let's use the same predictors in a correlation plot with Expected Goals percentage. Does Fenwick have the strongest correlation with xGoal percentage? -->

<!-- ```{r hockey ggplot 6} -->
<!-- xGoal_stats <- nhl_data_all %>% select(shotsOnGoalFor, corsiPercentage, fenwickPercentage, xGoalsPercentage) -->
<!-- ggpairs(xGoal_stats, title="Correlation plot of expected goals and relevant predictors, NHL 2021-22") -->
<!-- ``` -->


<!-- ## Rules of Basketball -->

<!-- [Link to YouTube video describing basketball rules](https://www.youtube.com/watch?v=wYjp2zoqQrs) -->

<!-- ## Basic Basketball Statistics -->

<!-- * **Field Goal (FG)**: A made shot from either 2- or 3-point range. Free throws, worth 1 point, are not considered field goals. Field goal statistics often include attempts, makes, and percentage. -->

<!-- * **Free Throw (FT)**: After certain fouls, the clock stops and a player shoots an uncontested shot from the foul line. These free throws are worth 1 point each; like with field goals, FT statistics often include attempts, makes, and percentage. -->

<!-- * **Assists (AST)**: A player is credited with an assist if they pass the ball to a teammate and the teammate scores a field goal after zero or one dribbles. No more than one assist can be recorded per field goal. -->

<!-- * **Turnover (TO)**: A player or team can be charged with a turnover for an action or violation that ends their offensive possession before being able to attempt a field goal. For a player (especially a guard), TOs can be compared to assists using Assist:Turnover ratio. -->

<!-- * **Rebound (REB)**: The first player to gain control of the ball following a missed field goal is credited with a rebound. If the player is on the same team as the field goal shooter, it is an offensive rebound; otherwise, a defensive rebound. -->

<!-- * **Points per Possession (PPP)**: Divides a team's points by number of possessions to account for a team's pace. -->

<!-- ## Advanced Basketball Statistics -->

<!-- * **True Shooting Percentage (TS%)**: Unlike traditional shooting percentage, this statistic considers both field goals and free throws. It also gives more weight to shots that are worth more points.  -->

<!-- * **Win Shares**: Win shares give each player points for actions that contribute to a team's success. Win shares take into account a variety of offensive and defensive statistics, but can be calculated using different methods on different platforms. -->

<!-- * **Value Over Replacement Player (VORP)**: This is basketball's response to baseball's WAR. VORP is a rate statistic that estimates a player's offensive output as compared to an "average" player. -->

<!-- * **Player Efficiency Rating (PER)**: According to its creator John Hollinger,  -->
<!-- "The PER sums up all a player's positive accomplishments, subtracts the negative accomplishments, and returns a per-minute rating of a player's performance." This statistic rewards great offensive performance more than great defensive plays. -->

<!-- *References: https://www.basketball-reference.com/about/per.html, https://www.basketball-reference.com/about/ws.html, https://www.basketball-reference.com/about/glossary.html* -->


<!-- ## Basketball - Four Factors -->

<!-- Tibbles are a type of data frame supported by the `tidyverse` package. The following tibble contains data from a Mountain West tournament game played between the CSU and Wyoming women's basketball teams during the 2021-2022 season, which CSU won 51-38. [(Here's the link to the box score on the CSU athletics website.)](https://csurams.com/sports/womens-basketball/stats/2021-22/wyoming/boxscore/16095) -->

<!-- ```{r four factors} -->
<!-- library("tibble") -->

<!-- basketball_data <- tibble(team = c("CSU","WYO"), FG = c(14,15), FGA = c(48,60), THREEP = c(5,4), FT = c(10,4), FTA = c(14,4), ORB = c(2,14), DRB = c(31,30), TOV = c(5,12)); basketball_data -->
<!-- ``` -->

<!-- This tibble contains all the data needed to calculate the *Four Factors*. The Four Factors of a basketball game are statistics formulated by Dean Oliver, former Director of Quantitative Analysis for the Denver Nuggets (among other roles). These statistics are also promoted by sports data platforms like Hudl.com. -->

<!-- The Four Factors each have offensive and defensive versions; for this example, we'll focus on the offensive perspective. -->

<!-- The first is **Effective Field Goal Percentage**, commonly abbreviated eFG%. The formula is as follows: -->

<!-- $eFG\% = \frac{FG\ +\ 0.5(3P)}{FGA}$ -->

<!-- Secondly, **Turnover Percentage** (TOV%) is calculated as: -->

<!-- $TOV\% = \frac{TOV}{FGA\ +\ 0.44(FTA)\ +\ TOV}$ -->

<!-- Next, **Rebounding Percentage** (ORB%) is computed as: -->

<!-- $ORB\% = \frac{ORB}{ORB\ +\ Opponent\ DRB}$ -->

<!-- Finally, the **Free Throw Factor** is found using: -->

<!-- $FT\ factor = \frac{FT}{FGA}$ -->

<!-- Note: You do not have to know these formulas for the test. They are just used for this example. -->

<!-- Let's calculate the values of eFG%, TOV%, ORB%, and Free Throw Factor for both CSU and Wyoming and add them as new columns in the tibble using the `add_column` function. -->

<!-- ```{r four factors 2} -->
<!-- attach(basketball_data) -->
<!-- eFG <- round((FG + .5 * THREEP)/FGA, 3) * 100 -->
<!-- TOVPCT <- round(TOV / (FGA + .44 * FTA + TOV), 3) * 100 -->
<!-- ORBPCT <- round(c(ORB[1]/(ORB[1]+DRB[2]), ORB[2]/(ORB[2]+DRB[1])), 3) * 100 -->
<!-- FTFACTOR <- round(FT/FGA, 3) * 100 -->

<!-- basketball_data %>% add_column(eFG, TOVPCT, ORBPCT, FTFACTOR) -->
<!-- ``` -->

<!-- While this method does produce Four Factors data, it could be difficult to scale for calculating the same statistics for a sample of several games. In the next section, we will introduce an `R` package that aids in the calculation of Four Factors. -->

<!-- ## Basketball - BasketballAnalyzeR Four Factors -->

<!-- The authors of "Basketball Data Science With Applications in R" developed the `BasketballAnalyzeR` package to be used in conjunction with the book. `BasketballAnalyzeR` includes built-in datasets from the 2017-18 NBA season and provides many functions for analyzing and plotting basketball data. One such function is `fourfactors`, which offers a simpler way to perform a four factors analysis. -->

<!-- ```{r basketballanalyzer 1, message=FALSE, warning=FALSE} -->
<!-- library("BasketballAnalyzeR") -->

<!-- teams <- c("DEN", "CLE", "GSW") #Nuggets, Cavaliers, Warriors -->
<!-- team_data <- which(Tadd$team %in% teams) -->
<!-- four_factors_teams <- fourfactors(Tbox[team_data, ], Obox[team_data, ]) -->
<!-- print(four_factors_teams) -->
<!-- ``` -->

<!-- This is a much simpler and neater way to calculate Four Factors. -->

In the 2017-18 season, the Warriors and Cavaliers met in the NBA Finals, while the Nuggets just missed the playoffs. It would be expected that the two Finals teams would have higher values for the Four Factors. While this is mostly true, for which of the offensive Four Factors did the Nuggets have a better value than the Cavaliers and Warriors?

A: Factor 3 (rebounding percentage), since 25.66 > 21.05 and 20.06. The only Factor in which the Nuggets were the best on the defensive side was also rebounding percentage.

<!-- ## Basketball - BasketballAnalyzeR Shot Charts -->

<!-- The `BasketballAnalyzeR` package includes shot location data for all players for the 2017-18 NBA season and has a function called `shotchart` that allows for the plotting of shot data. -->

<!-- Let's plot shot location data for Nikola Jokic. First, the coordinates must be transformed so that the point (0,0) is located at the corner of the court; the original coordinates place the origin at the center of the hoop. -->

<!-- ```{r basketballanalyzer 2} -->
<!-- PbP <- PbPmanipulation(PbP.BDB) -->

<!-- jokic_data <- subset(PbP, player == "Nikola Jokic") -->
<!-- jokic_data$xx <- jokic_data$original_x/10 #transformation -->
<!-- jokic_data$yy <- jokic_data$original_y/10 - 41.75 #transformation -->
<!-- ``` -->

<!-- `BasketballAnalyzeR` supports three types of density visualizations within `shotchart`, one being `density-polygons`. Since `shotchart` is a `ggplot` object, a chart title can be added using `ggtitle`. -->

<!-- ```{r basketballanalyzer 3} -->
<!-- shotchart(data = jokic_data, x="xx", y="yy", type="density-polygons") + ggtitle("Nikola Jokic Shot Data, 2017-18") -->
<!-- ``` -->

<!-- It seems most shots attempts from Jokic were in the paint; this is hardly surprising, since he plays the center position. Here's the same chart with `density-hexbin`: -->

<!-- ```{r basketballanalyzer 4} -->
<!-- shotchart(data = jokic_data, x="xx", y="yy", type="density-hexbin") + ggtitle("Nikola Jokic Shot Data, 2017-18") -->
<!-- ``` -->

<!-- The same chart with `density-raster`: -->

<!-- ```{r basketballanalyzer 5} -->
<!-- shotchart(data = jokic_data, x="xx", y="yy", type="density-raster") + ggtitle("Nikola Jokic Shot Data, 2017-18") -->
<!-- ``` -->

<!-- Within the `shotchart` function, setting `scatter=TRUE` overlays the shots on the chart. Point size and transparency can also be customized. -->

<!-- ```{r basketballanalyzer 6} -->
<!-- shotchart(data = jokic_data, x="xx", y="yy", type="density-raster", scatter=TRUE) + ggtitle("Nikola Jokic Shot Data, 2017-18") -->
<!-- ``` -->

<!-- Let's now compare shot charts of Nikola Jokic, Steph Curry, and Lebron James. This group of players includes one member of each team for which we calculated Four Factors. -->

<!-- ```{r basketballanalyzer 7} -->
<!-- curry_data <- subset(PbP, player == "Stephen Curry") -->
<!-- curry_data$xx <- curry_data$original_x/10 #transformation -->
<!-- curry_data$yy <- curry_data$original_y/10 - 41.75 #transformation -->

<!-- james_data <- subset(PbP, player == "LeBron James") -->
<!-- james_data$xx <- james_data$original_x/10 #transformation -->
<!-- james_data$yy <- james_data$original_y/10 - 41.75 #transformation -->

<!-- shotchart(data = jokic_data, x="xx", y="yy", type="density-raster") + ggtitle("Nikola Jokic Shot Data, 2017-18") -->
<!-- shotchart(data = curry_data, x="xx", y="yy", type="density-raster") + ggtitle("Steph Curry Shot Data, 2017-18") -->
<!-- shotchart(data = james_data, x="xx", y="yy", type="density-raster") + ggtitle("Lebron James Shot Data, 2017-18") -->
<!-- ``` -->

<!-- Q: Of the three players (Jokic, Curry, James), which took the highest percentage of their three-point shots from the right side of the court (when facing the basket)? -->

<!-- A: Nikola Jokic shot almost all of his attempts from the right side. Steph Curry took many shots from beyond the arc and tended toward the left side, while James was split between the right side and the center. -->

<!-- Now, let's focus on Steph Curry's shooting. The following chart splits the court into zones based on angle and distance from the basket. The color in each zone represents the average length of the play leading up to that shot among shots taken in that zone. -->

<!-- ```{r basketballanalyzer 8} -->
<!-- shotchart(data = curry_data, x="xx", y="yy", z="playlength", type="sectors", num.sect = 7, scatter=TRUE, pt.alpha=.3) + ggtitle("Steph Curry Shot Data, 2017-18") -->
<!-- ``` -->

<!-- Q: In general, did Steph Curry tend to shoot closer to the basket during plays of a longer duration or a shorter duration? -->

<!-- A: There is not a perfect correlation, but it seems that two-point field goals were attempted more often during longer plays, while shots taken outside the three-point arc were taken during plays of a shorter duration.  -->

<!-- *References: https://rdrr.io/cran/BasketballAnalyzeR/; Basketball Data Science (Zuccolotto and Manisera, 2020)* -->

<!-- ## Law of Total Probability - Hockey -->

<!-- Over the course of a season, a hockey player scored a goal 30% of the time during a home game, and $P(player\ scores\ |\ away\ game) = .18$. Assume all games are either home or away. -->

<!-- Q: What is the probability the player scored a goal in any game if there were an equal number of home and away games? -->

<!-- A: $P(score) = P(score|home)P(home) + P(score|away)P(away) = .3(.5) + .18(.5) = .24$ -->

<!-- Q: What is the probability the player scored a goal in any game if there were twice as many home games as away games? -->

<!-- A: $P(score) = P(score|home)P(home) + P(score|away)P(away) = .3(\frac{2}{3}) + .18(\frac{1}{3}) = .26$ -->

<!-- Q: What is the probability the player scored a goal in any game if the ratio of home games to away games is 2:3? -->

<!-- A: $P(score) = P(score|home)P(home) + P(score|away)P(away) = .3(\frac{2}{5}) + .18(\frac{3}{5}) = .228$ -->

<!-- ## Law of Total Probability - Baseball -->

<!-- You work in the front office of a professional baseball club and have just learned that a certain prospect hits .200 against left-handed pitchers and .400 against right-handed pitchers (their overall batting average is unknown). The general manager of the team overhears you talking about the .400 statistic of the player and becomes very exited that they have the chance to draft a .400 hitter. What would you say to caution the GM that the player might not be a remarkable hitter? -->

<!-- A: We don't know the proportion of the player's at-bats that came against left-handed pitchers versus right-handed pitchers. If we want to know the player's batting average unconditional on the type of pitcher they are facing, we have to adjust $P(hit\ |\ left-handed\ pitcher)$ by $P(left-handed\ pitcher)$ and $P(hit\ |\ right-handed\ pitcher)$ by $P(right-handed\ pitcher)$ before adding them to determine $P(hit)$. For example, if 90% of the player's at-bats were against left-handed pitchers, then their overall batting average is a pedestrian .220. -->

<!-- *Other possible issues: low sample size of player's at-bats, the fact that pro pitchers will be harder to hit against than non-pros* -->


<!-- ## Sets and Conditional Probability -->

<!-- 100 sports fans in Colorado were polled and it was found that 64 had attended either a Denver Nuggets or Colorado Avalanche game at Ball Arena (formerly Pepsi Center). 34 people had seen only a Nuggets game, while 17 had seen both a Nuggets and an Avalanche game. -->

<!-- Q: How many people saw an Avalanche game but not a Nuggets game? -->

<!-- A: 64 - 34 - 17 = 13 -->

<!-- Q: What is the probability that a randomly selected person in the poll had been to a Nuggets game? -->

<!-- A: (34 + 17) / 100 = .51 -->

<!-- Q: What is the probability that a randomly selected person that had been to a game at Ball Arena had been to a Nuggets game? -->

<!-- A: (34 + 17) / 64 = .797 -->

<!-- Q: What is the probability that a randomly selected person had been to a Nuggets game given they had been to an Avalanche game? -->

<!-- A: 17 / (17 + 13) = .567 -->


<!-- ## Binomial Probability -->

<!-- Two baseball teams are playing a 4-game series. The home team has a .65 probability of winning each game, and the away team a .35 probability. Assume each game is independent. -->

<!-- *I used baseball in this example because it's the sport that most often has 4-game series, but it could easily be replaced by another sport.* -->

<!-- Find the following probabilities. -->

<!-- (a) The road team wins exactly 1 game. -->

<!-- $\binom{4}{1}\ .65^3\ .35^1 = \binom{4}{3}\ .65^3\ .35^1 \approx .384$ -->

<!-- ```{r} -->
<!-- dbinom(1, 4, .35) -->
<!-- dbinom(3, 4, .65) -->
<!-- ``` -->

<!-- (b) The home team wins exactly 2 games. -->

<!-- $\binom{4}{2}\ .65^2\ .35^2 \approx .311$ -->

<!-- ```{r} -->
<!-- dbinom(2, 4, .65) -->
<!-- dbinom(2, 4, .35) -->
<!-- ``` -->

<!-- (c) The road team wins at least 2 games. -->

<!-- $\binom{4}{2}\ .65^2\ .35^2 + \binom{4}{3}\ .65^1\ .35^3 + .35^4 = 1 - [.65^4 + \binom{4}{1}\ .65^3\ .35^1]  \approx .437$ -->

<!-- ```{r} -->
<!-- pbinom(1.9, 4, .35, lower.tail=F) -->
<!-- pbinom(2, 4, .65, lower.tail=T) -->
<!-- ``` -->

<!-- (d) The series ends in a sweep. -->

<!-- $.65^4 + .35^4 \approx .194$ -->

<!-- ```{r} -->
<!-- dbinom(4, 4, .65) + dbinom(4, 4, .35) -->
<!-- .65^4 + .35^4 -->
<!-- ``` -->



<!-- ## Binomial Coefficient Symmetry -->

<!-- Playoff series for a certain sports league are played as a best-of-seven series, with one team hosting four games and the opposing team hosing three. An executive for the league wishes to know the number of ways the home and away games can be assigned. (One such combination is A-A-B-B-A-B-A, the format used by the NBA and NHL for their best-of-seven series.) What is the total number of combinations? -->

<!-- Answer: Since there are a fixed number of games (seven) and a fixed number of games that must be given to the lower-seeded team (four), there are $\binom{7}{4} = \frac{7!}{4!\ \cdot\ (7-4)!} = 35$ ways to create a home-away pattern for the seven-game series. -->

<!-- However, instead of thinking about the number of ways to assign the games to the team that gets four home games, what if we thought about the number of ways to assign games to the team that gets three home games? -->

<!-- That would be $\binom{7}{3}$. We can use the `choose` command in `R` to find this quantity. -->

<!-- ```{r bin} -->
<!-- choose(7,3) -->
<!-- ``` -->

<!-- It turns out that this binomial coefficient is also equal to 35. -->

<!-- Theorem: $\binom{n}{k} = \binom{n}{n-k}$ -->

<!-- $\binom{n}{k} = \frac{n!}{k!\ \cdot\ (n-k)!}$ -->

<!-- $\binom{n}{n-k} = \frac{n!}{(n-k)!\ \cdot\ (n-(n-k))!} = \frac{n!}{(n-k)!\ \cdot\ k!} = \binom{n}{k}$ -->


<!-- ## Binomials and Multinomials -->

<!-- Suppose we are curious about probabilities regarding the results of a soccer team’s next five games. -->

<!-- Wait!!! A soccer game has three possible outcomes (win, lose, draw)! We can’t use the binomial distribution, since it limits us to two possible outcomes! -->

<!-- It depends. If we are interested in the probability that a soccer team wins 2 of their next 5 games, we can use the binomial distribution. We can create the following partition of the sample space of outcomes: $(Win)$ and $(Win^C)$, where the second set includes both losing and drawing. -->

<!-- Then, the formula would be represented as: -->

<!-- $\binom{5}{2}\ P(Win)^2\  P(Win^C)^{(5-2)}$ -->

<!-- If we are interested in the probability of the team winning two of the next five games, drawing two, and losing one, we cannot use the binomial theorem. That involves three outcomes, and would be represented as a multinomial. <!--- assuming regularity conditions such as independence. --> -->

## Multinomial Distribution - Baseball

The following PMF displays the number of possible bases for a hit in baseball against the frequency of that type of hit for Hank Aaron, who is third all-time in Major League baseball hits with 3,771.

```{r multinomial, include=F}
Number_bases <- c(1:4)
Hit_Frequency <- c(2294, 624, 98, 755)
Hit_Proportion <- round(Hit_Frequency / sum(Hit_Frequency), 5)

Hank_Aaron_PMF <- data.frame(Number_bases, Hit_Proportion); Hank_Aaron_PMF
```

| Number of Bases, $x$  | 1 | 2 | 3 | 4 |
|:---------------------:|:-:|:-:|:-:|:-:|
| Probability, $P(X=x)$ | 0.61 | 0.16 | 0.03 | 0.20 |

Q: What is the probability that a randomly selected subset of 5 hits from Hank Aaron includes three singles, one double, and one home run?

<!-- note: assumes sampling with replacement. Without replacement would be a multivariate hypergeometric and would slightly increase the probability.  -->

A: From the Multinomial PMF, this is $P(X_1 = 3, X_2 = 1, X_3 = 0, X_4 = 1) = \frac{5!}{3!\ \cdot 1!\ \cdot 0!\ \cdot 1!}\ (.61)^3\cdot\ (.16)^1\cdot\ (.03)^0\cdot\ (.20)^1 \approx .145$

The probability can also be found in `R` using the `dmultinom` function.

```{r multinomial 2}
dmultinom(x = c(3, 1, 0, 1), prob = c(.61, .16, .03, .20))
```

<!-- ## Geometric (First Success) RVs -->

<!-- Caution: Some references parameterize the Geometric distribution based on the number of failures before the first success, rather than the trial on which the first success occurs. This changes the PMF, mean, and variance, so be careful. -->

<!-- ```{r} -->
<!-- set.seed(2020) -->
<!-- geometric <- rgeom(1e4, 1/10) -->
<!-- head(geometric, 20) -->
<!-- ``` -->

<!-- Some of the values were 0, which could not happen if `R` was considering the number of the trial on which the first success occurred. You can add 1 to the values given by `R` to arrive at the First Success distribution. -->

<!-- ```{r} -->
<!-- first_success <- geometric + 1 -->
<!-- head(first_success, 20) -->
<!-- mean(first_success) -->
<!-- ``` -->
<!-- The mean of this sample of variables is 10.11, which is close to the expected mean of $\frac{1}{p} = 10$. -->

<!-- ## Geometric Distribution - Hockey -->

<!-- Suppose the number of shots needed by a hockey team in order to score their first goal, $X$, is modeled by a Geometric($\frac{1}{10}$) random variable. -->

<!-- Q: What is the probability that it takes more than 3 shots to score the first goal? -->

<!-- A: $P(X > 3) = P(X = 4) + P(X = 5) + P(X = 6) +\ ...$ -->

<!-- This is an infinite series, so let's use the Complement Rule. -->

<!-- $P(X > 3) = 1 - P(X \leq 3) = 1 - \sum_{i=1}^{3}P(X = i) = 1 - [\frac{1}{10} + (\frac{9}{10})^1(\frac{1}{10}) + (\frac{9}{10})^2(\frac{1}{10})] = .729$ -->

<!-- Q: Why might the assumptions for the Geometric distribution make it an imperfect model for this situation? -->

<!-- A: Geometric random variables assume a constant probability for each trial (in this case, shots on goal). In hockey, the probability of a shot hitting the back of the net changes based on its location, velocity, positioning of defensive players, whether the team is on the power play, etc. -->

<!-- ## Poisson Distribution - Soccer -->

<!-- During the 2021 Major League Soccer season, the Colorado Rapids scored 51 goals in 34 games on their way to a first-place finish in the Western Conference regular season standings. -->

<!-- The team scored $\frac{51}{34} = 1.5$ goals per game. Let's model the distribution of Rapids goals using a Poisson(1.5) random variable that we'll call Y. -->

<!-- Q: Which is more likely: Y taking on the value 0 or Y taking on the value 2? -->

<!-- We can plot the PMF of Y to check visually. -->

<!-- ```{r soccer poisson} -->
<!-- ggplot(transform(data.frame(x=c(0:8)), y=dpois(x, lambda = 1.5)), aes(x, y)) + geom_bar(stat="identity") + labs(x="Value", y="Frequency", title="Probability mass function of Poisson(1.5) random variable") -->
<!-- ``` -->

<!-- A: From the plot, $P(Y = 2) > P(Y = 0)$. -->

<!-- Q: What is the exact probability that this Poisson(1.5) random variable takes on the value 2? -->

<!-- A: We can find this quantity using the `dpois` command. -->

<!-- ```{r soccer poisson 2} -->
<!-- dpois(x=2, lambda=1.5) -->
<!-- ``` -->
<!-- The same quantity can be obtained by plugging the numbers into the Poisson PMF. -->

<!-- $\frac{1.5^2 e^{-1.5}}{2!} = 0.251$ -->

<!-- Q: Let's switch it around. What is the probability that a Poisson(2) random variable takes on the value 1.5? -->

<!-- A: 0. Poisson random variables cannot take on values other than nonnegative integers. Remember, Poisson random variables are often used for count data. -->


<!-- Let's check whether using a Poisson distribution was appropriate by comparing it to the actual 2021 Colorado Rapids match results. -->

<!-- <!-- https://www.espn.com/soccer/team/results/_/id/184/season/2021 --> -->

<!-- ```{r soccer poisson 3} -->
<!-- library("kableExtra") -->

<!-- goals <- c(0:4, "5 or more") -->
<!-- actual_frequency <- c(6, 14, 7, 6, 0, 1) -->
<!-- actual_proportion <- actual_frequency / sum(actual_frequency) -->
<!-- expected_proportion <- c(dpois(0:4, lambda=1.5), ppois(4, lambda=1.5, lower.tail=FALSE)) -->
<!-- expected_frequency <- round(expected_proportion * 34, 1) -->

<!-- rapids.data <- data.frame(goals, actual_frequency, actual_proportion, expected_frequency, expected_proportion) -->

<!-- rapids.data %>% -->
<!-- kbl() %>% -->
<!-- kable_styling() -->
<!-- ``` -->

<!-- *answers may vary for following questions* -->

<!-- Q: What differences do you notice between the actual results and the expected values based on the Poisson random variable? -->

<!-- A: There were fewer games in which the Rapids scored 4 or more goals than the model would indicate, yet the Rapids were shut out less often than the model would indicate. -->

<!-- Q: Even if the true population distribution of 2021 Rapids goals was truly a Poisson(1.5) random variable, why might the actual distribution of their goals differ from the probability mass function? -->

<!-- A: 34 is a relatively small sample size; random variables may not coincide with their expected values for finite sample sizes. -->

<!-- Q: What are the advantages of using the Poisson distribution to model Major League soccer goals? What are the disadvantages? -->

<!-- A: Poisson random variables can take on the natural numbers (including zero), which aligns with the number of goals that can be scored in a match. One disadvantage is that it is possible for a Poisson to take on values that are not realistic for the situation, such as double-digit integers or higher. Only one game in MLS history has had a team score more than seven goals in a game. However, when $\lambda$ is small (such as 1.5), these extreme values are relatively unlikely. -->



## Multiple Probability Distributions - Basketball

Suppose the number of points scored by a basketball player follows a Poisson(12) random variable, the number of rebounds by a Poisson(7) distribution, and the number of assists by a random variable that is equally likely to take on any of the integers from 2 to 11, inclusive. Let points, rebounds, and assists be independent of each other.

<!-- Q: What is the probability that this player records a points, rebounds, assists triple-double in a game? -->

<!-- A: $P(Triple\ Double) = P(Points \geq 10\ \cap\ Rebounds \geq 10\ \cap\ Assists \geq 10)$ -->

<!-- ```{r basketball probability 1} -->
<!-- ppois(9, lambda = 12, lower.tail=F) -->
<!-- ``` -->

<!-- $P(Points \geq 10) = P(Poisson(12) \geq 10) \approx .758$ -->

<!-- ```{r basketball probability 2} -->
<!-- ppois(9, lambda = 7, lower.tail=F) -->
<!-- ``` -->

<!-- $P(Rebounds \geq 10) = P(Poisson(7) \geq 10) \approx .170$ -->

<!-- $P(Assists \geq 10) = P(Discrete\ Uniform(2, 11) \geq 10) = .2$ -->

<!-- Since the events are independent, we can multiply their probabilities. The probability of the player scoring the triple-double is $(.758)(.170)(.2) = .0257$. -->

<!-- Q: Your friend offers you 4 to 1 that the player will not record a triple-double in their next 10 games. With the knowledge that the athlete's performance in a game is unaffected by performances in previous games, would you take the bet? -->

<!-- A: $P(no\ triple\ double) = 1 - .0257 = .9743$, so $P(no\ triple\ double\ in\ next\ 10\ games) = (.9743)^{10} = .771$ -->

<!-- The odds of no triple-double are $\frac{.771}{1-.771} = 3.37$, so the bet of no triple-double at 4 to 1 odds is favorable. -->

<!-- ## Basketball Scenario -->

<!-- You are the coach of a basketball team that is down two points with one second remaining in the fourth quarter. During a timeout, you are considering the best play to call for your team. The first option is a three-point shot attempt, which you estimate has a 30% chance of succeeding. The second option is a two-point shot attempt, which has a 50% chance of making the field goal, a 30% chance of missing it and ending the game, and a 20% chance the shooter will miss but be fouled, in which case the shooter's free throw success will follow a $Bin(2, .8)$ random variable. Finally, you estimate that your team's probability of winning the game in overtime is .45. -->

<!-- Assume the above situations are exhaustive (i.e., the other team will not get another possession, no fouls will be called before the ball is put in play, lightning will not hit the arena and postpone the game, etc.). Which of the two plays should you call to maximize the win probability for your team? -->

<!-- A: The probability of winning the game with the three-point shot attempt is .3. If the two-point shot attempt is called for, there is a .5 probability of making the field goal and a (.2)(.8)(.8) = .128 probability that the foul is called and both free throws are made. Thus, the total probability of scoring two points and sending the game to overtime is .628. Then, the probability of winning the game in OT after tying it in regulation is (.628)(.45) = .2828. This is less than .3, so shooting the three-pointer is the option that maximizes the win probability, given these situational probabilities. -->

<!-- Q: What is the minimum estimated overtime win probability to make calling for the two-point play the better option? -->

<!-- A: $P(score\ 2\ points\ in\ regulation) \cdot P(win\ in\ OT) > P(win\ in\ regulation)$\ -->
<!-- $.628 \cdot P(win\ in\ OT) > .3$\ -->
<!-- $P(win\ in\ OT) > .478$ -->

<!-- ## Expectation - Baseball -->

<!-- The expectation of a discrete random variable is a weighted average. The "weights" are the probabilities of the possible values of the variable. -->

<!-- Consider the following table, which shows the number of career hits by type for the all-time Major League Baseball leader in total bases, Hank Aaron. -->

<!-- <!-- https://www.baseball-reference.com/players/a/aaronha01.shtml --> -->

<!-- ```{r expectation baseball, echo=F} -->
<!-- Hit_type <- c("Single", "Double", "Triple", "Home Run") -->
<!-- Number_bases <- c(1:4) -->
<!-- Hit_Frequency <- c(2294, 624, 98, 755) -->
<!-- Hit_Proportion <- round(Hit_Frequency / sum(Hit_Frequency), 8) -->

<!-- Hank_Aaron_Hits <- data.frame(Hit_type, Number_bases, Hit_Frequency, Hit_Proportion) -->

<!-- Hank_Aaron_Hits %>% -->
<!-- kbl() %>% -->
<!-- kable_styling() -->
<!-- ``` -->

<!-- The expected number of bases for a Hank Aaron hit is the sum of the number of bases attained for each hit multiplied by the relative frequency of the occurrence of that type of hit. -->

<!-- $1 \cdot \frac{2294}{3771} + 2 \cdot \frac{624}{3771} + 3 \cdot \frac{98}{3771} + 4 \cdot \frac{755}{3771} = 1.18181$ -->

<!-- This is the same process that is occurring whenever we calculate the expectation of any discrete random variable. Recall the formula for expectation is $E[X] = \sum_{x \in \Omega}\ x \cdot p(x)$. Each value in the sample space is "adjusted" by the probability of that value, then the sum of all values in $\Omega$ is taken to arrive at the weighted average, or expected value, of the random variable. -->


<!-- ## A few reminders/tips for simulation, and a basic example -->

<!-- The number of regulation goals scored in a game by Hockey Team A, $X$, is a Poisson(4) random variable, and the same for Hockey Team B, $Y$, is a Poisson(3.2) random variable. -->

<!-- A statistician is interested in the probability that Team A defeats Team B in regulation. This is $P(X > Y)$, which is difficult to calculate manually. However, using simulation, we can straightforwardly obtain an accurate estimation of this quantity. -->

<!-- There are many built-in functions in `R` that allow users to generate realizations from common probability distributions (`rnorm`, `rbinom`, `rexp`, etc.) Let's use the `rpois` function to simulate the appropriate variables, remembering to set a seed so that our results are easily replicable. -->

<!-- ```{r hockey poisson simulation 1} -->
<!-- set.seed(2022) -->

<!-- nReps <- 1e4 -->

<!-- team_A_goals <- rpois(n = nReps, lambda = 4) -->
<!-- team_B_goals <- rpois(n = nReps, lambda = 3.2) -->
<!-- ``` -->

<!-- Now, to find $P(X > Y)$, we can use the following line of code: -->

<!-- ```{r hockey poisson simulation 2} -->
<!-- mean(team_A_goals > team_B_goals) -->
<!-- ``` -->

<!-- Why does this work? First, operations to vectors are executed elementwise, meaning that `R` compares `team_A_goals[1]` to `team_B_goals[1]`, then `team_A_goals[2]` to `team_B_goals[2]`, and so on. Second, logical operators are stored as zeroes (when the condition is false) and ones (when the condition is true). The mean of a vector of zeroes and ones is the proportion of ones, which is the frequency of the logical statement being true. In our simulation, it was 0.5415. The true value is 0.5427, meaning that the simulation was quite accurate. -->

<!-- These tips will help you be more efficient when performing simulation tasks in `R`. -->

<!-- ```{r hockey poisson simulation 3, include=FALSE} -->
<!-- #computation for finding the theoretical value of P(X>Y) to within at least 7 decimal points. There's probably an easier way to do this. -->

<!-- v <- rep(NA, 50) -->

<!-- for (j in 1:50) { -->
<!--   v[j] <- ppois(j-1, lambda = 4, lower.tail=F) * dpois(j-1, lambda = 3.2) -->
<!-- } -->
<!-- sum(v) -->
<!-- ``` -->

<!-- ## Streak Simulation - Basketball -->

<!-- Suppose an NBA team is in the middle of a rebuild and has a 25% probability of winning each of its games in the following 82-game season. -->

<!-- Q: What is the probability that the team will go on at least one winning streak of four or more games over the course of the 82-game season? -->

<!-- A: We can simulate a season for the team, find the longest winning streak in that season, and store it in a vector. After repeating that process 10,000 times, we can then find the proportion of the values in that vector that are greater than or equal to 4. -->

<!-- ```{r basketball streaks simulation} -->
<!-- set.seed(2022) -->

<!-- nReps <- 1e4 -->
<!-- longest_streak <- rep(NA, nReps) -->

<!-- for (i in 1:nReps) { -->
<!--   game_results <- rbinom(size = 1, n = 82, prob = .25) # 1=win, 0=loss -->
<!--   streaks <- rle(game_results) -->
<!--   longest_streak[i] <- max(streaks$lengths[streaks$values==1]) -->
<!-- } -->

<!-- table(longest_streak) -->
<!-- mean(longest_streak >= 4) -->
<!-- ``` -->

<!-- <!-- https://stackoverflow.com/questions/28731582/maximum-consecutive-repeats-in-a-vector-in-r --> -->

<!-- The team had a 4+ game winning streak in about 20% of the simulations.  -->


<!-- # Properties of the Linear Correlation Coefficient -->

<!-- Pearson's correlation coefficient is unaffected by changes in location (adding a constant) or scale (multiplying by a constant) of data. To illustrate this concept, let's consider the correlation between the number of points scored by the winning team of the Super Bowl and 1) the number of the Super Bowl {1, 2, 3, ... , 56} and 2) the year of the Super Bowl {1967, 1968, 1969, ... , 2022}. -->

```{r correlation properties, include=FALSE}
Super_Bowl_Number <- c(1:56)
Super_Bowl_Year <- c(1967:2022)
Winning_Team_Points <- c(35,33,16,23,16,24,14,24,16,21,32,27,25,31,27,26,27,38,38,46,39,42,20,55,20,37,52,30,49,27,35,31,34,23,34,20,48,32,24,21,29,17,27,31,31,21,34,43,28,24,34,41,13,31,31,23)

Super_Bowl_Data <- data.frame(Super_Bowl_Number, Super_Bowl_Year, Winning_Team_Points)
```

```{r correlation properties 2, echo=FALSE}
ggplot(Super_Bowl_Data, aes(x=Super_Bowl_Number, y=Winning_Team_Points)) +
    geom_point() +
    geom_line(color="red") +
    ggtitle("Super Bowl Number and Points Scored by Winning Team")

ggplot(Super_Bowl_Data, aes(x=Super_Bowl_Year, y=Winning_Team_Points)) +
    geom_point() +
    geom_line(color="red") +
    ggtitle("Super Bowl Year and Points Scored by Winning Team")
```

```{r correlation properties 3}
cor(Super_Bowl_Number, Winning_Team_Points)
cor(Super_Bowl_Year, Winning_Team_Points)
```

<!-- The only difference between the two plots is that the x-axis is shifted. Since the change is essentially a linear (location) transformation, the correlations are identical. -->

<!-- Note: The correlations were equal only because there has been a one-to-one relationship between Super Bowl number and year since 1967. Major League Baseball has had years in which no World Series was held (1904 - boycott, 1994 - players' strike); thus, the same example applied to the MLB would produce slightly different correlation coefficients. -->

<!-- Q: The correlation coefficient is positive. What does this mean in terms of the trend of number of points scored by Super Bowl winners over time? -->

<!-- A: Winners of more recent Super Bowls are more likely to have scored more points than winners of older Super Bowls. -->

<!-- In a similar way to the example above, the correlation between a basketball player's career field goal attempts and three-pointers made and the correlation between the same player's career FGA and points scored from three-point field goals are exactly the same, due to the scale invariance property of correlation. -->

<!-- <!-- Note to Aaron: This could be a good homework exercise. Find data or simulate career FGA and 3PM for basketball players, then show that the two correlations described above are equal. --> -->


<!-- # Confounding Example -->

<!-- A statistician hypothesizes that an NBA team's draft position can be used to predict the number of regular-season wins for the team for the upcoming season. To investigate the relationship, they decide to estimate the following simple linear regression model: -->

<!-- $Wins\ = \beta_0 + \beta_1(Draft\ Position) + \epsilon_i$ -->

<!-- Using 2021-22 NBA data, the estimated intercept and slope for the model are: -->

<!-- ```{r confounding example, echo=FALSE} -->
<!-- NBA_Data <- read.csv("C:/Users/Fam/Downloads/NBA Draft and Win Data.csv") -->

<!-- #Note: Here, draft position is defined as the position the team "earns" in the drafting order based on their previous season performance, lottery draw, etc. This is not the same as draft pick number(s), as those are affected by trades and other factors. -->

<!-- lm(Wins_2021_2022 ~ Draft_Position_2021, data=NBA_Data) -->
<!-- ``` -->

<!-- Interpret the estimated slope in this model. -->

<!-- A: For every one-unit increase in draft position, the expected number of wins for the team increases by 0.6928. -->

<!-- After looking at the results of the model and the interpretation for the slope, one might identify that having a higher draft position is associated with more wins in the following season and conclude that teams with high draft picks should trade for lower draft picks in order to have a higher expected number of games won in the following season. What is the flaw in this reasoning? -->

<!-- <!-- possible answer: Teams receive higher draft positions when they are performing  worse in the previous season. Thus, it is expected that high draft picks will go to worse teams, and one high draft pick usually doesn't completely turn a team around in one season. From a statistical perspective, this SLR model does not take into account any confounding variables. --> -->

<!-- $~$ <!-- for whitespace --> -->

<!-- $~$ -->

<!-- $~$ -->

<!-- Recognizing that the SLR model may lead to erroneous lines of logic, the statistician decides to tackle the confounding by adding the wins for the team in the previous season to the model as a covariate. The theoretical model being estimated is now as follows: -->

<!-- $Wins\ = \beta_0 + \beta_1(Draft\ Position) + \beta_2(Previous\ Season\ Wins) + \epsilon_i$ -->

<!-- ```{r confounding example 2, echo=FALSE} -->
<!-- #Note: The 2021-22 season had an 82 game regular season, while the 2020-21 season had 72. (There hasn't been two consecutive seasons with equal numbers of GP since 2017-18, 2018-19). I considered having win percentage as the response variable, but ultimately decided to scale 2020-21 wins to an 82-game season and go from there. -->

<!-- lm(Wins_2021_2022 ~ Draft_Position_2021 + Wins_2020_2021_Scaled, data=NBA_Data) -->
<!-- ``` -->

<!-- When controlling for an NBA team's wins in the previous season, the sign for the slope coefficient for draft position flips to a negative. This suggests that for teams with the same winning percentage in the previous season, a lower draft position is associated with more wins in the following season. The results of this model, which included an important confounding variable, aligned with the conventional wisdom that lower draft positions are helpful for a team's future success.  -->

<!-- <!-- Note: I don't want to interpret the coefficients too in-depth in this second model since the two predictors are strongly correlated (high multicollinearity). The important concept here is the sign change of the draft position variable slope. --> -->

<!-- <!-- Homework problem idea: Use these models and the predict() function in R to simulate win totals for an NBA season. Include random noise in the model by picking a residual variance sigms and adding rnorm(0, sigma) to each observation. Then check to see if the totsl number of wins is close to (82/2)*30 = 1230 and see how the variance and histogram of the simulated win totals compare to actual NBA season results. Can also test the sensitivity of the simulation to different values of residual variance. --> -->

<!-- *Data: https://en.wikipedia.org/wiki/2021_NBA_draft, https://www.nba.com/standings?GroupBy=conf&Season=2020-21&Section=overall* -->


## Betting Strategy Expectations

A sports bettor wishes to wager on as many as 4 upcoming games; the bets are all uncorrelated. The bettor is considering between a traditional strategy of betting \$1 on each game and a martingale strategy where \$1 is bet on the first game amd the bet is doubled for the next game if the previous bet is not won, stopping once one bet is won. If any bet is won, the bettor is returned twice their wager; if the bet is lost, the bettor forfeits their wager. Suppose the house advantage is 4% and the bettor has no special skill (i.e., is no better than flipping a coin) at picking the games. 

Q: Calculate the probability the bettor wins their first wager.

A: $\frac{1 - 0.04}{2} = 0.48$

Find the expected value of the traditional betting strategy.\
*Hint: Find the expected value of one bet and then use the homogeneity property of expectation ($E[cX] = cE[X]$) to find the expected value of four bets.*

$E[Traditional\ bet] = (1)(.48) + (-1)(.52) = -0.04$\
$-0.04 * 4 = -0.16$

Find the expected value of the martingale betting strategy.\
*Hint: Consider two possible cases: one where a bet is won, and one where all four bets are lost.*

$E[Martingale] = E[Martingale\ |\ a\ bet\ is\ won] + E[Martingale\ |\ all\ four\ bets\ lost] = (1)(1 - .52^4) + (-15)(.52^4) = -0.17$

Q: Compare the expected values of each strategy. By observation, which strategy has the higher variance? What does this say about the martingale strategy and the tradeoff between probability and expectation in gambling?

A: The expected values are essentially equal, but the variance of the martingale strategy is much higher. While the probability of making a profit is higher using the martingale, the overall expectation is lower since there is about a 7% "risk of ruin" (i.e., losing all four bets).

Q: Which probability distribution has its right tail represented in the "risk of ruin" term in the martingale expectation?

A: Geometric(p = .48).

Extra Credit/Advanced: Generalize the martingale strategy expectation for $n$ possible uncorrelated bets, initial wager amount $w$, and constant probability of success $p$.\
*Hint: Consider the fact that $\sum_{i=0}^{n-1} 2^i = 2^n - 1$ for $n \geq 1$.*

$E[Martingale] = (w)(1 - (1-p)^n) + (-w)(2^n - 1)(1-p)^n$


## Negative Binomial Example

<!-- I might have p and 1-p backwards from the notation in the book -->

In 2022, the Colorado Mammoth won the National Lacrosse League championship. Over the course of the season, defenseman Tim Edwards took 363 faceoffs and won 187. Assume $\frac{187}{363} = .516$ is Edwards' true faceoff win percentage.

Q: Identify the appropriate distribution and parameters. What is the probability that Edwards' 5th faceoff win will come on his 9th attempt?

A: $X \sim NegBin(5,\ .516)$. For the 5th success to occur on the 9th trial (that is, for $X$ to equal 4), the first 8 attempts must contain 4 wins and 4 losses; they can be in any order. The expression for this probability is:

$\binom{8}{4} (.516)^4(1-.516)^4$

Then, the 9th faceoff attempt must be a success. The probability of success for that trial can be multiplied to the previous expression because of independence.

$\binom{8}{4} (.516)^4(1-.516)^4 \cdot (.516) = \binom{8}{4} (.516)^5(1-.516)^4 = .141$

As always, we will check our answer using the corresponding probability function in `R`.

```{r negbin}
dnbinom(4, 5, .516)
```

(Optional/Save for homework) Q: Provide an expression for the probability that it will take more than 15 faceoffs to observe Edwards' 7th faceoff win.

A: If it takes more than 15 trials to observe the seventh success, there were at least 9 failures on the way. Thus, we are interested in $P(X \geq 9)$. For all $k \geq 9$, there must be 6 successes and $k$ failures in the first $k + 6$ trials and then a success on the final trial. Since the sample space of a Negative Binomial experiment is uncountably infinite, this expression is an infinite sum.

$\sum_{k=9}^{\infty} \binom{k + 7 - 1}{k}(.516)^7(1-.516)^k$

```{r negbin 2}
pnbinom(8.5, 7, .516, lower.tail=FALSE)
```

If we want to avoid an infinite sum, the following expression is equivalent:

$1 - \sum_{k=0}^{8} \binom{k + 7 - 1}{k}(.516)^7(1-.516)^k$

```{r negbin 3}
1 - pnbinom(8, 7, .516, lower.tail=TRUE)
```

*Data: [coloradomammoth.com/team/stats/all-player-stats](https://coloradomammoth.com/team/stats/all-player-stats/)*

## Categorical Association - US Women's National Hockey Team

The US Women's National Hockey Team has reached the gold medal game in each of the past four Winter Olympic Games and won the gold in 2018. In the past three Olympiads, the team has played 19 games. Data on whether the team scored or allowed the first goal, split by wins and losses, is in the table below. Note that ties do not occur in Olympic hockey games.

```{r yules q, echo=FALSE}
uswnht <- data.frame(c(12, 2), c(2, 3))
colnames(uswnht) <- c("Won", "Lost")
rownames(uswnht) <- c("Scored First", "Did Not Score First")
uswnht %>% kt()
```

Q: What is the probability that the team won a game, conditional on the event they were the first team to score?

A: $\frac{12}{12+2} = .857$

Calculate the cross-product ratio for the above data.

$\frac{12 \cdot 3}{2 \cdot 2} = 9$

Q: If the cross-product ratio was lower (say, 3), would the knowledge of which team scored first be a more predictive estimate of the winning team?

A: No, knowing which team scored first would be less informative toward predicting the winning team since there would be more data points "in disagreement" rather than "in agreement".

Transform the cross-product ratio to Yule's Q.

$Q = \frac{9 - 1}{9 + 1} = 0.8$

Check:

$\frac{12 \cdot 3 - 2 \cdot 2}{12 \cdot 3 + 2 \cdot 2} = 0.8\  \checkmark$


## Categorical Association Homework Question

The Chicago Sky were the champions of the 2021 WNBA season. In this question, we'll investigate some of their regular season game-by-game stats.

Data: https://www.basketball-reference.com/wnba/teams/CHI/2021/gamelog/

1. Upload the dataset from Canvas. Note that in the column names, "Opp" stands for "Opponent". Create a column in the dataset with a 1 if the Sky won the game and a 0 otherwise.

```{r categorical association 1}
sky_data <- read.csv("C:/Users/Fam/Downloads/2021_Chicago_Sky_Data.csv")

sky_data <- sky_data %>%
  mutate(win = if_else(SkyPts > OppPts, 1, 0))
```

2. Basketball coaches often talk about "winning the rebound battle" as a key of winning a game. Create another column of an indicator variable for whether the Sky had more rebounds than their opponent.

```{r categorical association 2}
sky_data <- sky_data %>%
  mutate(rebBattle = if_else(SkyREB > OppREB, 1, 0))
```

3. Find the correlation between the two columns you just created. What does this mean about the relationship between winning the rebound battle and winning the game?

```{r categorical association 3}
cor(sky_data$win, sky_data$rebBattle)
```

Since $r > 0$, the data indicate that winning the rebound battle is associated with a greater probability of winning the game.

4. Use the `xtabs` function to create a 2x2 table of the frequencies of winning the rebound battle and winning the game. Save it as an object and print the table. Note: Ensure that the 1,1 cell of the table corresponds to *either* winning both the rebound battle and winning the game *or* losing both. If this is not the case, the answers to the following two questions will be different.

```{r categorical association 4}
sky_table <- xtabs(~win+rebBattle, data=sky_data)
sky_table
```

5. Find the cross-product ratio of the table. Does the ratio support your conclusion from part 3?

```{r categorical association 5}
ratio <- (sky_table[1,1] * sky_table[2,2]) / (sky_table[1,2] * sky_table[2,1])
ratio
```
Since $\alpha > 1$, the cross-product ratio indicates that winning the rebound battle is associated with a greater probability of winning the game, agreeing with the interpretation of the correlation.

6. Find Yule's Q for the table.

```{r categorical association 6}
q <- (ratio - 1) / (ratio + 1)
q
```

## Runs Test Homework Question

1. Using the same Chicago Sky dataset and the column you created, calculate the number of "runs" in the team's win/loss results.

```{r sky streaks}
sky_streaks <- rle(sky_data$win)
length(sky_streaks$lengths)
```

2. What is the expected number of runs for a team that finishes 16-16? What is the standard deviation?

$\mu = \frac{2 \cdot 16 \cdot 16}{32} + 1 = 17$

$\sigma = \sqrt{\frac{(17-1)(17-2)}{32-1}} = 2.78$

3. Find the z-score for the 2021 Sky results. Would the null hypothesis that wins and losses are randomly distributed throughout the season be rejected at the $\alpha = 0.05$ significance level? What does this say about the "streakiness" of the Sky?

$Z = \frac{13 - 17}{2.78} = -1.44$

At $\alpha = 0.05$, we would fail to reject the hypothesis that the wins and losses are randomly dispersed. We cannot conclude that the Sky were a "streaky" team in the 2021 regular season.

4. Is the expected number of runs higher for a team that finishes near the top or bottom of the WNBA standings, or for a team that finishes near .500 like the Sky? Explain.

One would expect that a team with a number of wins close to its number of losses would have more runs.