--- 
output:
  pdf_document: default
  html_document: default
---

## \# Ellie's Stuff

```{r}
library(dplyr)
library(ggplot2)
library(ggExtra)
library(ggrepel)
library(ggridges)
library(readr)
library(readxl)
library(openxlsx)

```

## To Be Added

### Soccer: EDA

Check out [this video](https://www.youtube.com/watch?v=cTCgU4Vz0rE&list=PLkBVjIpn_3P7dDtIMx4LuNtYf2Si9aobm) to learn how soccer is played, and [this video](https://www.youtube.com/watch?v=nT5qyrxoqsA) and/or [this video](https://www.youtube.com/watch?v=qg566N-eDWo) to learn about basic rules and regulations.

To begin, let's go over a couple of basic summary statistics specific to soccer that will be necessary to understand for the following examples.

-   **Shots (SH)** represent all shots taken by a team throughout the game. This is simply an attempt by a player to shoot the ball toward the net, even if they miss or the shot is saved (Rookie Road).

-   **Shots on Goal (SOG)** represent all shots that would have gone into the goal if not saved by a defender or goalkeeper (Rookie Road).

-   **Expected Goals (xG)** "indicates how many goals a team could have expected to score based on the quantity and quality of chances that they created in a match" (Tippett 2019, 4).

-   **Assist (A)** occur when a player passes the ball to someone, and the next shot results in a goal.

-   **Possession** refers to the percentage of time a team had control of the ball during a game.

These definitions come from www.rookieroad.com and "The Expected Goals Philosophy" by James Tippett.

To learn more about expected goals, check out [this video](https://www.youtube.com/watch?v=w7zPZsLGK18&list=PL9Az6mi38hv__afKnHR1AKjpcDIw_0qqT){.uri}.

### Bar Plot {.unnumbered}

Now that we have an understanding of some basic shooting statistics, let us go through some EDA examples. For this first example, we will need to install the "worldfootballR" package.

```{r}
library(worldfootballR)

```

Next we will look at some data specific to LaLiga, which is a soccer league in the men's top professional soccer division.

```{r}
# Get "Squad Standard Stats" Data
big5_2021_stats <- fb_big5_advanced_season_stats(season_end_year = 2021, 
                                                 stat_type = "standard", 
                                                 team_or_player = "team")
liga_2021_stats <- big5_2021_stats[which((big5_2021_stats$Comp == "La Liga")),]

```

```{r}
# Create visual for each team's goals per game
team_goals_viz <- 
  ggplot(data = liga_2021_stats[which(liga_2021_stats$Team_or_Opponent == "team"),], 
                         aes(x = Squad, y = Gls_Per)) +
  geom_bar(stat = "identity")
team_goals_viz

```

This plot is a good starting point, but still looks pretty messy. Let's add a title, change the axis titles, and rotate the axis labels so they are not overlapping over one another.

```{r}
team_goals_viz <- team_goals_viz +
  xlab("Team") +
  ylab("Goals Per Game") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Goals Per Game by Team")
team_goals_viz

```

This is already looking a lot better. Now, we will add the goals scored per game *against* each team. Why is this of interest? Well, at first glance, Barcelona seems like a pretty impressive team, as they score more goals per game than any other team in the league. However, what if they also have more goals scored against them than any other team in the league? This could be important context, so we will include it in the graph below.

```{r}
all_goals_viz <- ggplot(data = liga_2021_stats, aes(x = Squad, y = Gls_Per)) +
  geom_bar(stat = "identity", aes(fill = Team_or_Opponent), position = "stack") +
  xlab("Team") +
  ylab("Goals Per 90 Minutes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Goals Per Game by Team")
all_goals_viz

```

This is looking pretty good, but let's clean it up just a bit by changing the legend title and labels.

```{r}
all_goals_viz +
  scale_fill_discrete(name = "Team or Opponent", labels = c("Opponent","Team"))

```

What does this graph show us? Well, we are able to see the average number of goals scored for and against each team per game. It looks like Barcelona is scoring a lot more goals than they are letting be scored against them, while other teams like Valladolid tend to have a higher proportion of goals scored for the opposing team.

### Scatter Plot

In addition to simply knowing the average actual number of goals scored for and against each team per game, we may be interested in how this compares to the expected number of goals scored per game, as well.

```{r}
act_exp_viz <- ggplot(data = liga_2021_stats, aes(x = xG_Per, y = Gls_Per, label = Squad)) +
  geom_point() +
  scale_x_continuous(limits = c(0.75,2.25)) +
  scale_y_continuous(limits = c(0.75,2.25)) +
  ggtitle("Expected vs. Actual Goals Per Game") +
  xlab("Expected Goals Per Game") +
  ylab("Actual Goals Per Game") +
  geom_smooth(method = "lm", se = FALSE) +
  theme(aspect.ratio = 2/2)
ggMarginal(act_exp_viz, type = "density")

```

As you can see, we fit a line to the data. At first glance, it seems to have a positive slope slightly greater than 1. What does this mean in the scenario of actual and expected goals per game?

### Density Ridges Plot

At first glance, it seems that actual goals scored per game do not differ greatly from expected goals per game. Let us look at some density plots for actual and expected goals per game for five of the top teams in LaLiga over the last four seasons. These are the top five teams as of June 21st, 2022 on www.foxsports.com.

```{r}

```

```{r}
## NOT TO INCLUDE IN BOOK ##
# top_liga_2021_stats <- fb_big5_advanced_season_stats(season_end_year = 2021, stat_type = "standard", team_or_player = "team")
# top_liga_2020_stats <- fb_big5_advanced_season_stats(season_end_year = 2020, stat_type = "standard", team_or_player = "team")
# top_liga_2019_stats <- fb_big5_advanced_season_stats(season_end_year = 2019, stat_type = "standard", team_or_player = "team")
# top_liga_2018_stats <- fb_big5_advanced_season_stats(season_end_year = 2018, stat_type = "standard", team_or_player = "team")
# write_csv(top_liga_2021_stats,"~/Downloads/laliga21.csv")
# write_csv(top_liga_2020_stats,"~/Downloads/laliga20.csv")
# write_csv(top_liga_2019_stats,"~/Downloads/laliga19.csv")
# write_csv(top_liga_2018_stats,"~/Downloads/laliga18.csv")
```

```{r}
# Get "Squad Standard Stats" data for the last four seasons
top_liga_2021_stats <- read_csv("data/laliga21.csv")
top_liga_2020_stats <- read_csv("data/laliga20.csv")
top_liga_2019_stats <- read_csv("data/laliga19.csv")
top_liga_2018_stats <- read_csv("data/laliga18.csv")

top_liga_2021_stats <- top_liga_2021_stats[which(top_liga_2021_stats$Squad == "Real Madrid"|
                                           top_liga_2021_stats$Squad == "Villarreal"|
                                           top_liga_2021_stats$Squad == "Barcelona"|
                                           top_liga_2021_stats$Squad == "Levante"|
                                           top_liga_2021_stats$Squad == "Betis"),]
top_liga_2020_stats <- top_liga_2020_stats[which(top_liga_2020_stats$Squad == "Real Madrid"|
                                           top_liga_2020_stats$Squad == "Villarreal"|
                                           top_liga_2020_stats$Squad == "Barcelona"|
                                           top_liga_2020_stats$Squad == "Levante"|
                                           top_liga_2020_stats$Squad == "Betis"),]
top_liga_2019_stats <- top_liga_2019_stats[which(top_liga_2019_stats$Squad == "Real Madrid"|
                                           top_liga_2019_stats$Squad == "Villarreal"|
                                           top_liga_2019_stats$Squad == "Barcelona"|
                                           top_liga_2019_stats$Squad == "Levante"|
                                           top_liga_2019_stats$Squad == "Betis"),]
top_liga_2018_stats <- top_liga_2018_stats[which(top_liga_2018_stats$Squad == "Real Madrid"|
                                           top_liga_2018_stats$Squad == "Villarreal"|
                                           top_liga_2018_stats$Squad == "Barcelona"|
                                           top_liga_2018_stats$Squad == "Levante"|
                                           top_liga_2018_stats$Squad == "Betis"),]

# Combine all four seasons' data into one data frame
top_liga_stats <- rbind(top_liga_2018_stats, top_liga_2019_stats, top_liga_2020_stats, top_liga_2021_stats)

goals_act <- data.frame(top_liga_stats$Gls_Per[which(top_liga_stats$Team_or_Opponent == "team")])
goals_act$team <- top_liga_stats$Squad[which(top_liga_stats$Team_or_Opponent == "team")]
goals_act$exp_or_act <- "actual"
goals_act$year <- top_liga_stats$Season_End_Year[which(top_liga_stats$Team_or_Opponent == "team")]
colnames(goals_act)[1] <- "stats"
goals_exp <- data.frame(top_liga_stats$xG_Per[which(top_liga_stats$Team_or_Opponent == "team")])
goals_exp$team <- top_liga_stats$Squad[which(top_liga_stats$Team_or_Opponent == "team")]
goals_exp$exp_or_act <- "expected"
goals_exp$year <- top_liga_stats$Season_End_Year[which(top_liga_stats$Team_or_Opponent == "team")]
colnames(goals_exp)[1] <- "stats"
goals <- rbind(goals_act, goals_exp)

# Plot density ridges
ggplot(data = goals) +
  geom_density_ridges(aes(x = stats, y = team, fill = exp_or_act, color = exp_or_act),
                      alpha = 0.5, scale = 1) +
  scale_x_continuous(limits = c(0.75,2.75)) +
  scale_y_discrete(expand = expand_scale(add = c(0.2, 1))) +
  ggtitle("Expected vs. Actual Goals Per Game") +
  xlab("Goals Per Game") +
  ylab("Team") +
  scale_fill_cyclical(name = "Actual or Expected Goals",
                      labels = c("Actual","Expected"),
                      guide = "legend",
                      values = c("#FF0000A0", "#A0A0A0A0")) +
  scale_color_cyclical(name = "Actual or Expected Goals",
                       labels = c("Actual","Expected"),
                       guide = "legend",
                       values = c("#FF0000A0", "#A0A0A0A0"))

```

Let us break down exactly what this visual is showing us. We are looking at the density of average expected and actual goals per game for the top five teams in LaLiga, over the last four seasons (with the last season ending in 2021). We can see that Barcelona is typically scoring more goals than what is expected of them, as the density of actual goals is condensed around higher goal numbers than the density of expected goals. Villarreal, however, is performing just as well as what is expected of them based on expected and actual goals scored.

### Volleyball: EDA

Check out [this video](https://www.youtube.com/watch?v=9g7nYQv-kPM) to learn about the basics of volleyball.

To begin, let's go over some basic volleyball statistics. The following definitions come from www.rookieroad.com.

-   A **Service Ace (SA)** occurs when a player's serve touches the ground on the other team's side without being touched by a player on that side.

-   A **Kill (K)** occurs when a player gets the ball over the net without it being returned by the opponent.

-   An **Assist (AST)** is a pass made directly before a player makes a kill.

-   **Hitting Percentage (PCT)** is the number of attempted kills (minus errors) divided by the total number of kill attempts. This helps determine how well a player or team is succeeding at their kill attempts.

For Volleyball EDA, we will be using CSU Women's Volleyball data from the last five seasons.

```{r include=FALSE}
# Load CSU Women's Volleyball Data
csu_vb <- read_csv("data/csu_volleyball.csv")
colnames(csu_vb)[3] <- "W_L"

```

Let's look at a scatter plot of hitting percentage and the number of digs. While no conclusions can be drawn from such a plot, it can give us some insight into relationships worthy of further analysis. Before creating the plot using the code below, think about what you might expect the outcome to be. The following data comes from [this website](https://csurams.com/sports/womens-volleyball/stats).

### Scatter Plot

```{r}
# Digs, Hitting Percentage, Win/Lose
dig_pct_viz <- ggplot(data = csu_vb, aes(x = DIG, y = PCT, color = W_L)) +
  geom_point()
dig_pct_viz

```

Let's change the axis titles, legend title, and add a main title.

```{r}
dig_pct_viz +
  labs(title = "Wins and Losses by Number of Digs and Hitting Percentage",
       x = "Number of Digs (DIG)", y = "Hitting Percentage (PCT)",
       color = "Win or Loss")

```

What can we learn from this visual? Well, we can see that there is a weak linear relationship between the number of digs and hitting percentage. To an extent, hitting percentage decreases as the number of digs increases. Why is this the case? Maybe if a team has a really high hitting percentage, this means that the opposing team does not have as many opportunities to attack the other team offensively, reducing the number of opportunities for digs. It also seems that while wins and losses are somewhat evenly spread across the number of digs, there is a more clear cutoff for hitting percentage. It seems that the majority of wins are associated with a hitting percentage of at least 0.2, while the majority of losses are associated with a hitting percentage of less than 0.3.

### Box Plot

Now let's take a closer look at the distribution of hitting percentage and digs for wins and losses. To do this, we will create box plots for each statistic.

```{r}
pct_viz <- ggplot(data = csu_vb, aes(x = PCT, y = W_L)) +
  geom_boxplot()
pct_viz

dig_viz <- ggplot(data = csu_vb, aes(x = DIG, y = W_L)) +
  geom_boxplot()
dig_viz

```

Let's modify these plots to make them more complete and visually appealing.

```{r}
pct_viz +
  labs(title = "Hitting Percentage for Wins and Losses",
       x = "Hitting Percentage (PCT)",
       y = "Win or Loss") +
  geom_boxplot(fill = "slateblue", alpha = 0.2)

dig_viz +
  labs(title = "Number of Digs for Wins and Losses",
       x = "Number of Digs (DIG)",
       y = "Win or Loss") +
  geom_boxplot(fill = "slateblue", alpha = 0.2)

```

Box plots allow us to isolate each statistic (number of kills and hitting percentage) so we can more clearly determine the center and spread of each between wins and losses.

### Soccer: Probability

### Poisson Distribution {.unnumbered}

```{r include=FALSE}
season_18_19 <- read_csv("data/laliga_home_goals.csv")

```

```{r}
set.seed(1234)

home_goals <- season_18_19 %>%
  count(FTHG)

lambda <- (0*82 + 1*127 + 2*111 + 3*45 + 4*10 + 5*4 + 8*1)/(82 + 127 + 111 + 45 + 10 + 4 + 1)

# lambda is close to 1.5

values <- rpois(380, lambda = 1.5)
df <- data_frame(x = c(0:6), values = table(values)) 

ggplot() +
  geom_bar(data = home_goals, aes(x = FTHG, y = n), stat = "identity") +
  geom_smooth(aes(x = c(0:6), y = df$values), se = FALSE) +
  xlab("Number of Goals Scored by Home Team") +
  ylab("Frequency") +
  ggtitle("LaLiga Home Goals and Poisson Distribution")

```

Using the Poisson(1.5) distribution, answer the following question: What is the probability a home team scores 3 goals?

Recall the probability mass function for the Poisson distribution: $p(x) = \frac{e^{-x}\lambda^x}{x!}$

We need to find $p(x = 3)$.

$$
p(x = 3) = \frac{e^{-3}1.5^3}{3!} \approx 2.8\%
$$

[Insert another example here]

### Volleyball: Probability

### Conditional Probability {.unnumbered}

The CSU Women's Volleyball Team has a double-header coming up. The probability that they will win the first game is 80%. The probability that they will win both games is 55%. What is the probability that they will win the second game given that they win the first game?

This is a case of conditional probability. Recall that $P(B | A) = \frac{P(A \cap B)}{P(A)}$. Let $A$ be the event that the CSU Women's Volleyball Team wins the first game. Let $B$ be the event that they win the second game. We can solve for $P(B | A)$ as follows:

$$
P(B | A) = \frac{0.55}{0.80} = 0.6875
$$

Therefore, the probability of the CSU Women's Volleyball Team winning the second game given that they win the first game is 68.75%.

### Autocorrelation (from Analytic Methods in Sports pg. 134)

```{r}
# Combining Data, Do Not Include in Book

setwd("/Users/elliemartinez/Course\ Development/sports_data/")
file.list <- list.files(pattern='*xlsx')
df.list <- lapply(file.list, read_excel)
all_teams <- bind_rows(df.list, .id = "id")
write.csv(all_teams, "/Users/elliemartinez/Course\ Development/nfl_win_pct.csv")

```

```{r}
# Problem from pg. 134 of Analytic Methods (not complete yet as of 7/21)

nfl_win_pct <- read_csv("data/nfl_win_pct.csv")

all_teams$win_pct <- df$W/(df$W + df$L)

autocorr <- function(x) {
  return (acf(x, 1, plot = FALSE)$acf[2])
}

autocorrs <- aggregate(df$win_pct, list(df$id), FUN = autocorr)
mean_autocorrs <- mean(autocorrs$x)

```

### Rank Correlation (from Analytic Methods in Sports pg. 129)

```{r}

```

### Random Forests

```{r}

```

------------------------------------------------------------------------
